---
title: "Jordy_Data_Analysis"
author: "Jordan Von Eggers"
date: "2023-10-18"
output: html_document
editor_options: 
  chunk_output_type: console
---

# 0. Set up on Beartooth 
Request resources, move to directory, and load modules for R
```{bash}
ssh jvonegge@beartooth.arcc.uwyo.edu
cd /project/lakecolor/data_analysis

salloc --mem=110GB --nodes=1 --cpus-per-task=1 --account=microbiome --time=2:00:00

module load arcc/1.0  gcc/12.2.0 
module load r/4.2.2
R
```

```{r}
require(tidyverse)
#color for manually created legend
bg.fui = tibble(
  ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
  ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
  color = c(
  "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
  "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
  "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04")
)
```



# 1. Histogram of temporal lake size
```{r}
load("2023-10-19_TEMPORAL_Limnosat_May_Oct_Lagos_Filtered.RData")


pdf(paste0(Sys.Date(),"_TEMPORAL_HistogramLakeSize.pdf"), height=8,width=7)
hist(temporal$lake_waterarea_ha, xlim=c(0,1000),breaks=100000)
dev.off()


uniq_lakes <-temporal[,names(temporal)%in%c("lake_nhdid","lake_waterarea_ha")]
uniq_lakes<-unique(uniq_lakes)

uniq_lakes_small<-uniq_lakes[uniq_lakes$lake_waterarea_ha<11,]


pdf(paste0("/gscratch/jvonegge/LakeColor/",Sys.Date(),"_TEMPORAL_HistogramLakeSize_SmallLakes.pdf"), height=8,width=7)
hist(uniq_lakes_small$lake_waterarea_ha,ylab="Freq", xlab="Lake size (hectares)", main="Histogram of temporal dataset lakes < 11 ha")
dev.off()


uniq_lakes_med<-uniq_lakes[uniq_lakes$lake_waterarea_ha<100,]

pdf(paste0("/gscratch/jvonegge/LakeColor/",Sys.Date(),"_TEMPORAL_HistogramLakeSize_LakesLessThan100ha.pdf"), height=8,width=7)
hist(uniq_lakes_med$lake_waterarea_ha, ylab="Freq", xlab="Lake size (hectares)", main="Histogram of temporal dataset lakes < 100 ha")
dev.off()

rm(list=setdiff(ls(), "temporal"))

```

# 2. run linear models for temporal dataset
## a. run on supercomputer
```{r}
require(tidyverse)
load("../../data/2023-10-19_TEMPORAL_Limnosat_May_Oct_Lagos_Filtered.RData")

bg.fui = tibble(
        ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
        ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
        color = c(
                "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
                "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
                "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04")
)

bg.fui<-as.data.frame(bg.fui)

temporal_avg_lms <- temporal %>%
        group_by(lake_nhdid) %>%
        mutate(range_dwl_site=max(dWL)-min(dWL)) %>% 
        ungroup() %>%
        group_by(lake_nhdid,year) %>%
        mutate(avg_dwl_year=mean(dWL), 
               range_dwl_site_year= max(dWL)- min(dWL),
               obs_year=n(),
               obs_days_year=as.numeric(max(as.Date(date)) - min(as.Date(date)))) %>% ungroup() %>%
        group_by(lake_nhdid) %>%
        mutate(mean_range_dwl_site=mean(range_dwl_site_year)) %>% ungroup() %>%
        select(-date,-dWL,-brightness,-normalized_green,-Nir,-lake_year) %>% 
        distinct() %>%
        
        #linear models
        group_by(lake_nhdid) %>%
        mutate(lm_yint=summary(lm(avg_dwl_year~year))$coefficients[1,1],
               lm_pval=summary(lm(avg_dwl_year~year))$coefficients[2,4],
               lm_rsq=summary(lm(avg_dwl_year~year))$r.squared,
               lm_slope=summary(lm(avg_dwl_year~year))$coefficients[2,1],
               residuals=abs(lm(avg_dwl_year~year)$residuals),
               resid_yint=summary(lm(residuals~year))$coefficients[1,1], 
               resid_pval=summary(lm(residuals~year))$coefficients[2,4],
               resid_rsq=summary(lm(residuals~year))$r.squared,
               resid_slope=summary(lm(residuals~year))$coefficients[2,1]) %>% ungroup() %>%
        
        #adding in hex color
        mutate( col_hex = case_when(avg_dwl_year >=bg.fui[1,1] & avg_dwl_year <bg.fui[1,2] ~ bg.fui[1,3],
                                    avg_dwl_year >=bg.fui[2,1] & avg_dwl_year <bg.fui[2,2] ~bg.fui[2,3], 
                                    avg_dwl_year >=bg.fui[3,1] & avg_dwl_year <bg.fui[3,2] ~bg.fui[3,3]   , 
                                    avg_dwl_year >=bg.fui[4,1] & avg_dwl_year <bg.fui[4,2] ~bg.fui[4,3]   , 
                                    avg_dwl_year >=bg.fui[5,1] & avg_dwl_year <bg.fui[5,2] ~bg.fui[5,3]   , 
                                    avg_dwl_year >=bg.fui[6,1] & avg_dwl_year <bg.fui[6,2] ~bg.fui[6,3]   , 
                                    avg_dwl_year >=bg.fui[7,1] & avg_dwl_year <bg.fui[7,2] ~bg.fui[7,3]   , 
                                    avg_dwl_year >=bg.fui[8,1] & avg_dwl_year <bg.fui[8,2] ~bg.fui[8,3]   , 
                                    avg_dwl_year >=bg.fui[9,1] & avg_dwl_year <bg.fui[9,2] ~bg.fui[9,3]   , 
                                    avg_dwl_year >=bg.fui[10,1] & avg_dwl_year <bg.fui[10,2] ~bg.fui[10,3]   , 
                                    avg_dwl_year >=bg.fui[11,1] & avg_dwl_year <bg.fui[11,2] ~bg.fui[11,3]   , 
                                    avg_dwl_year >=bg.fui[12,1] & avg_dwl_year <bg.fui[12,2] ~bg.fui[12,3]   , 
                                    avg_dwl_year >=bg.fui[13,1] & avg_dwl_year <bg.fui[13,2] ~bg.fui[13,3]   , 
                                    avg_dwl_year >=bg.fui[14,1] & avg_dwl_year <bg.fui[14,2] ~bg.fui[14,3]   , 
                                    avg_dwl_year >=bg.fui[15,1] & avg_dwl_year <bg.fui[15,2] ~bg.fui[15,3]   , 
                                    avg_dwl_year >=bg.fui[16,1] & avg_dwl_year <bg.fui[16,2] ~bg.fui[16,3]   , 
                                    avg_dwl_year >=bg.fui[17,1] & avg_dwl_year <bg.fui[17,2] ~bg.fui[17,3]   , 
                                    avg_dwl_year >=bg.fui[18,1] & avg_dwl_year <bg.fui[18,2] ~bg.fui[18,3]   , 
                                    avg_dwl_year >=bg.fui[19,1] & avg_dwl_year <bg.fui[19,2] ~bg.fui[19,3]   , 
                                    avg_dwl_year >=bg.fui[20,1] & avg_dwl_year <bg.fui[20,2] ~bg.fui[20,3]   , 
                                    avg_dwl_year >=bg.fui[21,1] & avg_dwl_year <=bg.fui[21,2] ~bg.fui[21,3] , T~"black")) %>%
        
        #adding in category
        group_by(lake_nhdid) %>%
        mutate(dwl_prior2002=mean(avg_dwl_year[year <= 2002])) %>%
        ungroup() %>%
        
        mutate(trend_cat= case_when(
                lm_pval <= 0.05 & lm_slope >= 0 & dwl_prior2002 < 530 ~ 'Blue -> Greener',
                lm_pval <= 0.05 & lm_slope >= 0 & dwl_prior2002 > 530 ~ 'Intensifying Green/brown',
                lm_pval <= 0.05 & lm_slope <= 0 & dwl_prior2002 > 530 ~ 'Green -> Bluer',
                lm_pval <= 0.05 & lm_slope <= 0 & dwl_prior2002 < 530 ~ 'Intensifying Blue',
                lm_pval > 0.05 & dwl_prior2002 < 530 ~ 'No trend - Blue',
                lm_pval > 0.05 & dwl_prior2002 > 530 ~ 'No trend - Green/brown'
        ),
        trend_cat = factor(
                trend_cat,
                levels = c(
                        'No trend - Blue',
                        'No trend - Green/brown',
                        'Intensifying Blue',
                        'Green -> Bluer',
                        'Intensifying Green/brown',
                        'Blue -> Greener'
                )
        )
        )


write.csv(temporal_avg_lms,paste0(Sys.Date(),"_TEMPORAL_Avg_LMs.csv"))
```



run_2023-11-27_TEMPORAL_Avg_LMs.sh
```{bash}
#!/bin/bash
#SBATCH --job-name temporal_lms
#SBATCH --mem=100GB
#SBATCH --time=4-00:00:00
#SBATCH --cpus-per-task=1
#SBATCH --account=microbiome
#SBATCH --output=temporal_lms_%A.out

cd /project/lakecolor/data_analysis
module load arcc/1.0  gcc/12.2.0 
module load r/4.2.2

srun Rscript 2023-11-05_TEMPORAL_Avg_LMs.R
date
```

## 2a. Removing lakes that have dominant wavelengths outside of 470-590 for >5 observations. 28 lakes were removed from the analyses.
```{r}
temporal<-read.csv("2023-12-05_TEMPORAL_Avg_LMs.csv", header=T,row.names=1)

dWL_subset  <- temporal |>
  filter(avg_dwl_year > 590 | 
           avg_dwl_year < 470) |>
  group_by(lagoslakeid) |>
  count() |>
  filter(n > 5) # 28 lakes

delete_lakes <- dWL_subset$lagoslakeid

temporal <- temporal |>
  filter(!lagoslakeid %in% delete_lakes)

```


## b. figure out "black" points
```{r}
temporal<-read.csv("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2023-12-05_TEMPORAL_Avg_LMs.csv", header=T,row.names=1)

#black observations
black <- temporal %>% filter(col_hex=="black") %>% select(lake_nhdid,lake_lat_decdeg, lake_lon_decdeg,avg_dwl_year, year, col_hex) %>% distinct() %>% filter(avg_dwl_year > 590) %>% count(lake_ndhid)

```


## c. rename ecoregion
```{r}
temporal<-read.csv("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2023-12-05_TEMPORAL_Avg_LMs.csv", header=T,row.names=1)

temporal_uniq<-temporal %>%
        select(lake_nhdid, lake_lat_decdeg, lake_lon_decdeg, trend_color,trend_cat,epanutr_zoneid) %>%
        distinct()

cols <- c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99", # lt pink
  "palegreen2",
  "#FDBF6F", # lt orange
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
  "darkturquoise", "green1", "yellow4", "yellow3",
  "darkorange4", "brown","gray","khaki2","#CAB2D6")

temporal_uniq$ecoreg_color <- cols[as.factor(temporal_uniq$epanutr_zoneid )]

legend<-temporal_uniq %>%
        select(epanutr_zoneid ,ecoreg_color) %>%
        distinct() %>% arrange(epanutr_zoneid)

print(legend)

  epanutr_zoneid ecoreg_color
1      epanutr_1  "dodgerblue2"
2      epanutr_2      "#E31A1C"
3      epanutr_3       "green4"
4      epanutr_4      "#6A3D9A"
5      epanutr_5      "#FF7F00"
6      epanutr_6        "black"
7      epanutr_7        "gold1"
8      epanutr_8     "skyblue2"
9      epanutr_9      "#FB9A99"

par(mar=c(3,3,3,7), xpd=T)
plot(temporal_uniq$lake_lon_decdeg, temporal_uniq$lake_lat_decdeg, col= temporal_uniq$ecoreg_color, pch=16, cex=0.5, bty="n")
legend("right",legend=legend$epanutr_zoneid,col=legend$ecoreg_color, pch=16, pt.cex=3, inset = c(-0.2, 0.1))

temporal<-temporal %>% 
        mutate(ecoregion_name = case_when(epanutr_zoneid=="epanutr_1"~"Coastal Plains",
                                          epanutr_zoneid=="epanutr_2"~"Northern Appalachians",
                                          epanutr_zoneid=="epanutr_3"~"Northern Plains",
                                          epanutr_zoneid=="epanutr_4"~"Southern Appalachians",
                                          epanutr_zoneid=="epanutr_5"~"Southern Plains",
                                          epanutr_zoneid=="epanutr_6"~"Temperate Plains",
                                          epanutr_zoneid=="epanutr_7"~"Upper Midwest",
                                          epanutr_zoneid=="epanutr_8"~"Western Mountains",
                                          epanutr_zoneid=="epanutr_9"~"Xeric", T~"black"))
```

plot ecoregions
```{r}

cols <- rev(c(
  "dodgerblue2", "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black", "gold1",
  "skyblue2", "#FB9A99"))

temporal_uniq<-temporal %>%
        select(lake_nhdid, lake_lat_decdeg, lake_lon_decdeg, trend_color,trend_cat,epanutr_zoneid, ecoregion_name) %>%
        distinct()
temporal_uniq$ecoregion_name <- factor(temporal_uniq$ecoregion_name, levels=c("Xeric", "Western Mountains","Upper Midwest","Temperate Plains","Southern Plains","Southern Appalachians", "Northern Plains","Northern Appalachians" ,"Coastal Plains"))

temporal_uniq$ecoregion_color <- cols[temporal_uniq$ecoregion_name]

temporal_uniq$ecoregion_color <- factor(temporal_uniq$ecoregion_color, levels=cols)

temporal_uniq %>%
        select(ecoregion_color,ecoregion_name) %>%
        distinct() %>% arrange(ecoregion_name,levels(temporal_uniq$ecoregion_name)) %>% print()

  ecoregion_color        ecoregion_name
1         "#FB9A99"                 Xeric
2        "skyblue2"     Western Mountains
3           "gold1"         Upper Midwest
4           "black"      Temperate Plains
5         "#FF7F00"       Southern Plains
6         "#6A3D9A" Southern Appalachians
7         "green4"       Northern Plains
8         "#E31A1C" Northern Appalachians
9     "dodgerblue2"        Coastal Plains


eco<-ggplot(data = temporal_uniq, aes(x = lake_lon_decdeg, y = lake_lat_decdeg, color=ecoregion_color)) +
  geom_point(size = 0.7) +
  scale_color_manual(values = levels(temporal_uniq$ecoregion_color), 
                     labels = levels(temporal_uniq$ecoregion_name)) + 
        guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_void() +
  coord_fixed(1.3) +
labs(color = "Ecoregion") +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), fill = "transparent", color = "black", linewidth = 0.5) +
        theme(legend.position = "right",  # Change the legend position to the right
        legend.box.margin = margin(l = -20))




jpeg(paste0("Figures/",Sys.Date(),"_Temporal_EcoRegionsMap.jpg"),height=6,width=9, res=500, units="in")
eco
dev.off()
```

## d. plot trends on map
in ggplot
```{r}
require(ggplot2)
# Get state borders
states <- map_data("state")

temporal_uniq<-temporal %>%
        select(lake_nhdid, lake_lat_decdeg, lake_lon_decdeg, trend_cat) %>%
        distinct()
temporal_uniq$trend_cat <- factor(temporal_uniq$trend_cat, levels=c("Intensifying Blue","No trend - Blue","Green -> Bluer", "Blue -> Greener","No trend - Green/brown","Intensifying Green/brown"))


cols <- c("blue1", #Intensifying Blue,
            "#E0EEEE", #No trend - Blue
            "skyblue2",  # Green -> Bluer
          "#A2CD5A", # Blue -> Greener
    "#C1FFC1", #  No trend - Green/brown
  "green4") #Intensifying Green/brown

temporal_uniq$trend_color <- cols[temporal_uniq$trend_cat]

temporal_uniq$trend_color <- factor(temporal_uniq$trend_color, levels=cols)

trends<-temporal_uniq %>%
        select(trend_cat,trend_color) %>%
        distinct() %>% arrange(trend_cat,levels(temporal_uniq$trend_cat))

print(trends)

                 trend_cat trend_color
1        Intensifying Blue      "blue1"
2          No trend - Blue     "#E0EEEE"
3           Green -> Bluer    "skyblue2"
4          Blue -> Greener     "#A2CD5A"
5   No trend - Green/brown     "#C1FFC1"
6 Intensifying Green/brown      "green4"


map_trends<-ggplot(data = temporal_uniq, aes(x = lake_lon_decdeg, y = lake_lat_decdeg, color=trend_color)) +
  geom_point(size = 0.7) +
  scale_color_manual(values = levels(temporal_uniq$trend_color), 
                     labels = levels(temporal_uniq$trend_cat)) + 
        guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_void() +
  coord_fixed(1.3) +
labs(color = "Lake color trend") +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), fill = "transparent", color = "black", linewidth = 0.5) +
        theme(legend.position = "right",  # Change the legend position to the right
        legend.box.margin = margin(l = -20))


jpeg(paste0("Figures/",Sys.Date(),"_LakeColorTrends_EcoRegionsMap.jpg"),height=6,width=9, res=500, units="in")
map_trends
dev.off()
```
check in base R
```{r}
cols <- c("#A2CD5A", # Blue -> Greener
  "skyblue2",  # Green -> Bluer
  "blue1", #Intensifying Blue
  "green4", #Intensifying Green/brown
  "#E0EEEE", #No trend - Blue
  "#C1FFC1" #  No trend - Green/brown
)
temporal$trend_color <- cols[as.factor(temporal$trend_cat)]

trends<-temporal %>%
        select(trend_cat,trend_color) %>%
        distinct()
print(trends)

#                  trend_cat trend_color
# 1   No trend - Green/brown     "#C1FFC1"
# 2 Intensifying Green/brown      "green4"
# 3        Intensifying Blue      "blue1"
# 4           Green -> Bluer   "skyblue2"
# 5          No trend - Blue     "#E0EEEE"
# 6          Blue -> Greener     "#A2CD5A"

temporal_uniq<-temporal %>%
        select(lake_nhdid, lake_lat_decdeg, lake_lon_decdeg, trend_color,trend_cat) %>%
        distinct()

plot(temporal_uniq$lake_lon_decdeg, temporal_uniq$lake_lat_decdeg, col= temporal_uniq$trend_color, pch=16, cex=0.5)
```




## e. plot proportion by ecoregion
```{r}
temporal<-temporal %>% 
        mutate(ecoregion_name = case_when(epanutr_zoneid=="epanutr_1"~"Coastal Plains",
                                          epanutr_zoneid=="epanutr_2"~"Northern Appalachians",
                                          epanutr_zoneid=="epanutr_3"~"Northern Plains",
                                          epanutr_zoneid=="epanutr_4"~"Southern Appalachians",
                                          epanutr_zoneid=="epanutr_5"~"Southern Plains",
                                          epanutr_zoneid=="epanutr_6"~"Temperate Plains",
                                          epanutr_zoneid=="epanutr_7"~"Upper Midwest",
                                          epanutr_zoneid=="epanutr_8"~"Western Mountains",
                                          epanutr_zoneid=="epanutr_9"~"Xeric", T~"black"))


temporal_ecoreg <- temporal %>%
        select(lake_nhdid, trend_cat,trend_color,ecoregion_name) %>%
        group_by(ecoregion_name) %>%
        count(trend_cat)%>% 
  mutate(proportion = n / sum(n))


temporal_ecoreg$trend_cat <- factor(temporal_ecoreg$trend_cat, levels=c("Intensifying Blue","No trend - Blue","Green -> Bluer", "Blue -> Greener","No trend - Green/brown","Intensifying Green/brown"))


cols <- c("blue1", #Intensifying Blue,
            "#E0EEEE", #No trend - Blue
            "skyblue2",  # Green -> Bluer
          "#A2CD5A", # Blue -> Greener
    "#C1FFC1", #  No trend - Green/brown
  "green4") #Intensifying Green/brown

temporal_ecoreg$trend_color <- cols[temporal_ecoreg$trend_cat]

temporal_ecoreg$trend_color <- factor(temporal_ecoreg$trend_color, levels=cols)

temporal_ecoreg[,names(temporal_ecoreg)%in%c("trend_cat","trend_color")] %>%
        distinct() %>% arrange(trend_cat,levels(temporal_ecoreg$trend_cat))


  trend_cat                trend_color
  <fct>                    <fct>      
1 Intensifying Blue        "blue1"      
2 No trend - Blue          "#E0EEEE"    
3 Green -> Bluer           "skyblue2"   
4 Blue -> Greener          "#A2CD5A"    
5 No trend - Green/brown   "#C1FFC1"    
6 Intensifying Green/brown "green4" 



temporal_ecoreg$ecoregion_name <- factor(temporal_ecoreg$ecoregion_name, levels=c("Xeric", "Western Mountains","Upper Midwest","Temperate Plains","Southern Plains","Southern Appalachians", "Northern Plains","Northern Appalachians" ,"Coastal Plains"))

gplt<-ggplot(data = temporal_ecoreg,
       aes(
               x = ecoregion_name,
               y = proportion,
               group = trend_cat,
               col = trend_color
       )) + 
        scale_color_manual(values = levels(temporal_ecoreg$trend_color), 
                           labels = levels(temporal_ecoreg$trend_cat)) + 
        geom_line(linewidth = 2) + 
        geom_point(size = 3) + 
        theme_bw() + 
        labs(color = "Lake color trend") + 
        xlab(label = "Ecoregion") + 
        ylab(label = "Proportion") +
        theme(axis.text.x = element_text(angle = 45, hjust=1))

jpeg(paste0("Figures/",Sys.Date(),"_LakeColorTrend_ByEcoregion.jpg"),height=4,width=9, res=500, units="in")
gplt
dev.off()


```



# 3. Figure 1 
## a. histogram of dwl

```{r}

bg.fui = tibble(
        ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
        ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
        color = c(
                "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
                "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
                "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04"))

bg.fui<-as.data.frame(bg.fui)


cols<-as.data.frame(breaks_seq<-seq(470,590,by=5))
names(cols)<-"breaks"
cols<- cols %>%
mutate( col_hex = case_when(breaks >=bg.fui[1,1] & breaks <bg.fui[1,2] ~ bg.fui[1,3],
                                    breaks >=bg.fui[2,1] & breaks <bg.fui[2,2] ~bg.fui[2,3], 
                                    breaks >=bg.fui[3,1] & breaks <bg.fui[3,2] ~bg.fui[3,3]   , 
                                    breaks >=bg.fui[4,1] & breaks <bg.fui[4,2] ~bg.fui[4,3]   , 
                                    breaks >=bg.fui[5,1] & breaks <bg.fui[5,2] ~bg.fui[5,3]   , 
                                    breaks >=bg.fui[6,1] & breaks <bg.fui[6,2] ~bg.fui[6,3]   , 
                                    breaks >=bg.fui[7,1] & breaks <bg.fui[7,2] ~bg.fui[7,3]   , 
                                    breaks >=bg.fui[8,1] & breaks <bg.fui[8,2] ~bg.fui[8,3]   , 
                                    breaks >=bg.fui[9,1] & breaks <bg.fui[9,2] ~bg.fui[9,3]   , 
                                    breaks >=bg.fui[10,1] & breaks <bg.fui[10,2] ~bg.fui[10,3]   , 
                                    breaks >=bg.fui[11,1] & breaks <bg.fui[11,2] ~bg.fui[11,3]   , 
                                    breaks >=bg.fui[12,1] & breaks <bg.fui[12,2] ~bg.fui[12,3]   , 
                                    breaks >=bg.fui[13,1] & breaks <bg.fui[13,2] ~bg.fui[13,3]   , 
                                    breaks >=bg.fui[14,1] & breaks <bg.fui[14,2] ~bg.fui[14,3]   , 
                                    breaks >=bg.fui[15,1] & breaks <bg.fui[15,2] ~bg.fui[15,3]   , 
                                    breaks >=bg.fui[16,1] & breaks <bg.fui[16,2] ~bg.fui[16,3]   , 
                                    breaks >=bg.fui[17,1] & breaks <bg.fui[17,2] ~bg.fui[17,3]   , 
                                    breaks >=bg.fui[18,1] & breaks <bg.fui[18,2] ~bg.fui[18,3]   , 
                                    breaks >=bg.fui[19,1] & breaks <bg.fui[19,2] ~bg.fui[19,3]   , 
                                    breaks >=bg.fui[20,1] & breaks <bg.fui[20,2] ~bg.fui[20,3]   , 
                                    breaks >=bg.fui[21,1] & breaks <=bg.fui[21,2] ~bg.fui[21,3] , T~"black"))



png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/",Sys.Date(),"_Histogram_dwl_colored.png"), height=4, width=5, units="in", res=500)
par(mar=c(4,4,1,1))
hist(Y$avg_dwl_site, ylab="Number of lakes", xlab="Dominant wavelength (nm)",breaks=cols$breaks, col = cols$col_hex, main="", ylim=c(0,8000))
abline(v=530, col="black", lty="dashed",lwd=2)
dev.off()


```



## b. PCA
Did this in the super computer because was running SO slow on my computer

Read in climate data, average, and merge with the trend dataset, and then with the landcover dataset

Request resources, move to directory, and load modules for R
```{bash}
ssh jvonegge@beartooth.arcc.uwyo.edu
cd /project/lakecolor/data

salloc --mem=110GB --nodes=1 --cpus-per-task=1 --account=microbiome --time=2:00:00

module load arcc/1.0  gcc/12.2.0 
module load r/4.2.2
R
```
### i. compile trends, climate, landcover


```{r}
require(tidyverse)

# 1. read in climate data from PRISM
prism<-readRDS("prism_lakepoint_data.rds")
range(prism$year) # range of PRSIM data
#[1] "1984" "2022"

# average the mean temperature, the temperature range, rainfall, fo the 39 years of observation
prism_avg <- prism %>%
        group_by(lake_nhdid) %>%
        mutate(
                avg_temp_degC = mean(tmean),
                range_temp_degC = mean(tmax - tmin), # this is tmax and min within a year
                avg_ppt = mean(ppt),
                obs_year_prism = n()) %>%
        ungroup() %>%
        select(lake_nhdid,avg_temp_degC,range_temp_degC, avg_ppt,obs_year_prism) %>%
        distinct()
dim(prism_avg) # 145083 sites
table(prism_avg$obs_year_prism) # all have 39 years
summary(prism_avg) # 2 NAs..

# read in temporal trend data
trends<-read.csv("2023-12-11_TEMPORAL_Avg_LMs.csv", header=T, row.names=1)

trends$X<-NULL

# label lakes that ACTUALLY shifted from blue to green or green to blue 
trend_change <- trends %>%
        group_by(lake_nhdid) %>%
        mutate(dwl_after2002 = mean(avg_dwl_year[year > 2002])) %>%
        ungroup() |>
        mutate(
                trend_cat_switch = ifelse(
                        trend_cat %in% c('Blue -> Greener', 'Green -> Bluer') &
                                dwl_prior2002 > 530 &
                                dwl_after2002 < 530,
                        'Shift to blue',
                        ifelse(
                                trend_cat %in% c('Blue -> Greener', 'Green -> Bluer') &
                                        dwl_prior2002 < 530 &
                                        dwl_after2002 > 530,
                                'Shift to green',
                                trend_cat)))

trend_change %>%
        filter(trend_cat_switch %in%c('Shift to green','Shift to blue'))%>%
        select(lake_nhdid) %>%
        distinct() %>%
        nrow()
# [1] 5133 - same as what Linnea found in code for figure 3


trends_avg <- trend_change %>%
        group_by(lake_nhdid) %>%
        mutate(avg_obs_site = mean(obs_year),
               avg_obs_days_site = mean(obs_days_year),
               avg_dwl_site = mean(avg_dwl_year),
               avg_range_dwl_site = mean(range_dwl_site_year)) %>%
                       ungroup()%>%
        select(
                lake_nhdid,
                epanutr_zoneid,
                hu12_zoneid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                lake_islandperimeter_m,
                lake_connectivity_class,
                lake_connectivity_permanent,
                lake_totalperimeter_m,
                ws_area_ha,
                ws_lake_arearatio,
                range_dwl_site,
                mean_range_dwl_site,
                lm_yint,
                lm_pval,
                lm_rsq,
                lm_slope,
                resid_yint,
                resid_pval,
                resid_rsq,
                resid_slope,
                dwl_prior2002,
                trend_cat,
                trend_cat_switch,
                avg_obs_site,
                avg_obs_days_site,
                avg_dwl_site,
                avg_range_dwl_site
        )%>%
        distinct()
dim(trends_avg)


# join the trend data and prism data
trends_prism <- inner_join(trends_avg,prism_avg, by= join_by(lake_nhdid))
dim(trends_prism)
#[1] 81007    35

# read in landcover data, filter by lakes we have in trends, and average all columns
landcover <- readRDS("/project/lakecolor/data/archive/2023-09-29_lagos_landcover_data.rds")

landcover_avg <- landcover %>%
        filter(lake_nhdid %in% trends$lake_nhdid) %>%
        select(
                lake_nhdid,
                year,
                datacoveragepct,
                nlcd_barren31_pct,
                nlcd_cultcrop82_pct,
                nlcd_devhi24_pct,
                nlcd_devlow22_pct,
                nlcd_devmed23_pct,
                nlcd_devopen21_pct,
                nlcd_forcon42_pct,
                nlcd_fordec41_pct,
                nlcd_formix43_pct,
                nlcd_grass71_pct,
                nlcd_icesnow12_pct,
                nlcd_openwater11_pct,
                nlcd_past81_pct,
                nlcd_shrub52_pct,
                nlcd_wetemerg95_pct,
                nlcd_wetwood90_pct
        ) %>% 
        group_by(lake_nhdid) %>%
        mutate(
                landcover_first_year = min(year),
                landcover_last_year = max(year),
                obs_year_landcover = n(),
                across(starts_with("nlcd"), list(avg = ~ mean(.)), .names = "avg_{.col}")
        ) %>%
        ungroup() %>%
        select(
                lake_nhdid,
                landcover_first_year,
                landcover_last_year,
                obs_year_landcover,
                avg_nlcd_barren31_pct,
                avg_nlcd_cultcrop82_pct,
                avg_nlcd_devhi24_pct,
                avg_nlcd_devlow22_pct,
                avg_nlcd_devmed23_pct,
                avg_nlcd_devopen21_pct,
                avg_nlcd_forcon42_pct,
                avg_nlcd_fordec41_pct,
                avg_nlcd_formix43_pct,
                avg_nlcd_grass71_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_openwater11_pct,
                avg_nlcd_past81_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_wetemerg95_pct,
                avg_nlcd_wetwood90_pct
        ) %>% # get rid of year and data from individual years
        distinct() 

# combine all development landcover types
landcover_avg_sum <- landcover_avg %>%
        rowwise() %>%
        mutate(avg_nlcd_dev_pct = sum(
                c(
                        avg_nlcd_devlow22_pct,
                        avg_nlcd_devhi24_pct,
                       avg_nlcd_devmed23_pct,
                       avg_nlcd_devopen21_pct
                ))) 


#merge landcover with prism and lake color trends
trends_prism_landcover <- inner_join(trends_prism,landcover_avg_sum, by= join_by(lake_nhdid))
dim(trends_prism_landcover)
#[1] 81007    55



# select the variables you want to work on, but it will automatically grab date and lake_ndhid
X <- trends_prism_landcover %>%
        select(
                lake_nhdid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                lake_islandperimeter_m,
                ws_area_ha,
                avg_temp_degC,
                range_temp_degC,
                avg_ppt,
                avg_nlcd_barren31_pct,
                avg_nlcd_cultcrop82_pct,
                avg_nlcd_dev_pct,
                avg_nlcd_forcon42_pct,
                avg_nlcd_fordec41_pct,
                avg_nlcd_formix43_pct,
                avg_nlcd_grass71_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_openwater11_pct,
                avg_nlcd_past81_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_wetemerg95_pct,
                avg_nlcd_wetwood90_pct
        )
dim(X)
#[1] 81007    24

X <- X %>%
       column_to_rownames(var = "lake_nhdid") %>% distinct() %>% drop_na()
dim(X)
#[1] 80931    23

Y <- trends_prism_landcover %>%
               filter(lake_nhdid%in%rownames(X))%>%
        select(lake_nhdid,avg_dwl_site, trend_cat, trend_cat_switch, lm_slope)
 
Y <- Y %>%
       column_to_rownames(var = "lake_nhdid") %>% distinct() %>% drop_na()
dim(Y)
#[1] 80931     4

rm(list=ls()[! ls() %in% c("Y", "X")])

save.image(paste0(Sys.Date(),"_Trends_Prism_Landcover.Rdata"))
```
### - look at correlations of predictors
```{r}
cormat<-cor(landcover[18:33])

require(ggcorrplot)
pdf("Landcover_correlation.pdf",height=10,width=10)
ggcorrplot(cormat)
dev.off()

cormat<-cor(as.matrix(trends_avg[,c("lake_elevation_m","lake_lat_decdeg" , "lake_lon_decdeg","lake_waterarea_ha" , "lake_perimeter_m","lake_islandperimeter_m","lake_totalperimeter_m","ws_area_ha","ws_lake_arearatio")]),use="pairwise.complete.obs")

pdf("Lakevar_correlation.pdf",height=10,width=10)
ggcorrplot(cormat)
dev.off()
```

### load data
```{r}
load("../../../2024-03-07_Trends_Prism_Landcover.rdata")
```

### ii. run PCA
```{r}
#S <- cov(X)
R <- cor(X)
round(R,4)

pc <- princomp(X, cor=T)
spc <- summary(pc, loadings=T, cutoff=0)
lambda <- pc$sd^2

# look at correlation
png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor",Sys.Date(),"_correlation_vars_X_PCA.png"), height=10, width=10, units="in",res=400)
psych::pairs.panels(X,
             lm=T,
             method="pearson",
             density=T,
             hist.col = "green4",
             ellipses=F,
             ci=F)
dev.off()
```

### - decide how many PCs
```{r}
spc
plot(pc, type="lines", main = "")
```
Look at the summary and decide how many PCAs you want to keep based on
1. The proportion of variance (try to get up to 80%)
 - 8 components gets 0.682, pretty good, first component is 17 percent..
2. Eigenvalues in "Importance of components" - Standard deviation should be greater than 1
- eigenvalues are above one up until component 8
3. Where is "the elbow in the plot"
- there is a big elbow after the first component.
Decision - Keep 8 PCs

### - Interpret loadings using a cutoff of 0.3
```{r}
for (i in 1:8){
print(paste0("Principal component",i))
print(pc$loadings[which(abs(pc$loadings[,i])>0.3),i])
}
```
PC1: high elevation, higher ice and snow are lower longitude (in the east), lower precip
PC2: lower latitudes have higher temperatures and more precipitation
PC3: lake area, perimeter, island perimeter, and watershed area are all positively correlated
PC4: high forest mix, high grass, low shrub
PC5: low range in temperature more barren land, less deciduous forest, and more wooded wetlands
PC6: lower range in temperatures, higher cultivated crops, higher pasture, higher shrub, lower wetemerg
PC7: lower development, more pasture, more wooded wetlands
PC8: more cultivated crops, more conifer forests, less mixed forest, more open water, less shrubs


### iii. plot PCs
```{r}
bg.fui = tibble(
        ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
        ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
        color = paste0(c(
                "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
                "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
                "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04"),"60")
)

bg.fui<-as.data.frame(bg.fui)

       
Y_dwl<- Y %>%
        rownames_to_column() %>%
        rename(lake_nhdid=rowname) %>%
        select(lake_nhdid,avg_dwl_site)

pc.scores <- as.data.frame(pc$scores)
pc.scores<-pc.scores %>% 
        rownames_to_column() %>%
        rename(lake_nhdid=rowname) %>%
        inner_join(.,Y_dwl, by=join_by(lake_nhdid)) %>%
        mutate( col_hex = case_when(avg_dwl_site >=bg.fui[1,1] & avg_dwl_site <bg.fui[1,2] ~ bg.fui[1,3],
                                    avg_dwl_site >=bg.fui[2,1] & avg_dwl_site <bg.fui[2,2] ~bg.fui[2,3], 
                                    avg_dwl_site >=bg.fui[3,1] & avg_dwl_site <bg.fui[3,2] ~bg.fui[3,3]   , 
                                    avg_dwl_site >=bg.fui[4,1] & avg_dwl_site <bg.fui[4,2] ~bg.fui[4,3]   , 
                                    avg_dwl_site >=bg.fui[5,1] & avg_dwl_site <bg.fui[5,2] ~bg.fui[5,3]   , 
                                    avg_dwl_site >=bg.fui[6,1] & avg_dwl_site <bg.fui[6,2] ~bg.fui[6,3]   , 
                                    avg_dwl_site >=bg.fui[7,1] & avg_dwl_site <bg.fui[7,2] ~bg.fui[7,3]   , 
                                    avg_dwl_site >=bg.fui[8,1] & avg_dwl_site <bg.fui[8,2] ~bg.fui[8,3]   , 
                                    avg_dwl_site >=bg.fui[9,1] & avg_dwl_site <bg.fui[9,2] ~bg.fui[9,3]   , 
                                    avg_dwl_site >=bg.fui[10,1] & avg_dwl_site <bg.fui[10,2] ~bg.fui[10,3]   , 
                                    avg_dwl_site >=bg.fui[11,1] & avg_dwl_site <bg.fui[11,2] ~bg.fui[11,3]   , 
                                    avg_dwl_site >=bg.fui[12,1] & avg_dwl_site <bg.fui[12,2] ~bg.fui[12,3]   , 
                                    avg_dwl_site >=bg.fui[13,1] & avg_dwl_site <bg.fui[13,2] ~bg.fui[13,3]   , 
                                    avg_dwl_site >=bg.fui[14,1] & avg_dwl_site <bg.fui[14,2] ~bg.fui[14,3]   , 
                                    avg_dwl_site >=bg.fui[15,1] & avg_dwl_site <bg.fui[15,2] ~bg.fui[15,3]   , 
                                    avg_dwl_site >=bg.fui[16,1] & avg_dwl_site <bg.fui[16,2] ~bg.fui[16,3]   , 
                                    avg_dwl_site >=bg.fui[17,1] & avg_dwl_site <bg.fui[17,2] ~bg.fui[17,3]   , 
                                    avg_dwl_site >=bg.fui[18,1] & avg_dwl_site <bg.fui[18,2] ~bg.fui[18,3]   , 
                                    avg_dwl_site >=bg.fui[19,1] & avg_dwl_site <bg.fui[19,2] ~bg.fui[19,3]   , 
                                    avg_dwl_site >=bg.fui[20,1] & avg_dwl_site <bg.fui[20,2] ~bg.fui[20,3]   , 
                                    avg_dwl_site >=bg.fui[21,1] & avg_dwl_site <=bg.fui[21,2] ~bg.fui[21,3] , T~"black"))


# rename variable labels

rownames(pc$loadings)<- c("Elevation (m)" , "Latitude" , "Longitutde",
"Lake area (ha)" , "Lake perimeter (m)" , "Island perimeter (m)",
"W.S. area (ha)" , "Temp (C)" , "Range temp (C)",
"Precip (ppt)" , "Barren (%)" , "Cultcrop (%)",
"Dev (%)" , "Forcon (%)" , "Fordec (%)",
"Formix (%)" , "Grass (%)" , "Icesnow (%)",
"Openwater (%)" , "Past (%)" , "Shrub (%)",
"Wetemerg (%)" , "Wetwood (%)")





i=1
for(i in 1:7){
load <- pc$loadings[, c(i,i+1)]
load <- load[abs(load[,1])>0.3 | abs(load[,2])>0.3, ]

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/PCA/",Sys.Date(),"_PC",i,"_PC",i+1,"_SepScale.png"), width = 8, height=6, units = "in", res=400)
par(mar=c(5,5,1,1), xpd=T)
plot(pc.scores[,i+1], pc.scores[,i+2], pch = 19, col = pc.scores$col_hex, xlab=paste0("PC",i),ylab=paste0("PC",i+1))

# scaling factor for arrow length
scaling_factor <- c(11,28,14,9,11,9,7)[i]


# Add arrows for loadings of PC1 and PC2
arrows(0, 0, load[, 1] * scaling_factor, load[, 2] * scaling_factor, 
       angle = 20, length = 0.1, col = "black")


text(load[, 1] * scaling_factor, load[, 2] * scaling_factor,
     labels = rownames(load), pos = 3, col = "black", cex = 1.1)
dev.off()
}

rm(load)
rm(scaling_factor)
rm(bg.fui)
```


### iv. MLR PCA
Multiple linear regression using ALL PCs
```{r}
lake_color<- pc.scores %>% select(avg_dwl_site, paste0("Comp.",seq(1:23)))
(round(RY<-cor(lake_color)[,1][2:9],digits=2))
# Comp.1 Comp.2 Comp.3 Comp.4 Comp.5 Comp.6 Comp.7 Comp.8 
#   0.03   0.15   0.09  -0.65  -0.11  -0.01   0.12   0.02
```


component 4 has a moderate, negative correlation with lake color, all others are pretty weak.
PC4: high forest mix, high grass, low shrub
This makes sense because its the best separator in the plots
Component 2 has the second higest correlation at 0.14.. so maybe trying plots 4 and 2 together would be good, or component 7.

```{r}
lake_color<- pc.scores %>% select(avg_dwl_site, paste0("Comp.",seq(1:23)))

mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)

lake_color <- lake_color %>% select(avg_dwl_site, paste0("Comp.",seq(1:8)))

mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)

lake_color <- lake_color %>% select(avg_dwl_site,  paste0("Comp.",c(2,4,7)))

mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)


lake_color <- lake_color %>% select(avg_dwl_site,  paste0("Comp.",c(4,2,7)))

mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)
```



### v. MLR all preds
```{r}
X_N <- rownames_to_column(X, var = "rownames")
Y_N <- rownames_to_column(Y, var = "rownames")
lake_color <- inner_join(X_N,Y_N,by="rownames") %>% 
        select(-c(trend_cat,trend_cat_switch,lm_slope)) %>%
               column_to_rownames(var = "rownames") %>%
        select(avg_dwl_site, everything())
(RY<-cor(lake_color)[,1])
# avg_dwl_site         lake_elevation_m          lake_lat_decdeg 
#               1.00000000              -0.12522647              -0.26192961 
#          lake_lon_decdeg        lake_waterarea_ha         lake_perimeter_m 
#              -0.16193585              -0.02097072              -0.04523071 
#   lake_islandperimeter_m               ws_area_ha            avg_temp_degC 
#              -0.02251970               0.02006931               0.32718941 
#          range_temp_degC                  avg_ppt    avg_nlcd_barren31_pct 
#               0.23908082              -0.18216747              -0.07609383 
#  avg_nlcd_cultcrop82_pct         avg_nlcd_dev_pct    avg_nlcd_forcon42_pct 
#              -0.09074182              -0.01934530              -0.10529881 
#    avg_nlcd_fordec41_pct    avg_nlcd_formix43_pct     avg_nlcd_grass71_pct 
#              -0.23543373              -0.34409469              -0.37377726 
#   avg_nlcd_icesnow12_pct avg_nlcd_openwater11_pct      avg_nlcd_past81_pct 
#              -0.03993905               0.32601892               0.24336466 
#     avg_nlcd_shrub52_pct  avg_nlcd_wetemerg95_pct   avg_nlcd_wetwood90_pct 
#               0.37594538              -0.23793292               0.10507696

X_highcor <- names(sort(RY[abs(RY) > 0.3],decreasing = T))
```
Pretty weak correlation individually which supports using PCA

```{r}
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)
# r-adj: 0.4952

lake_color <- lake_color %>% select(-c(lake_waterarea_ha,lake_waterarea_ha,avg_nlcd_shrub52_pct))

mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)
# r-adj:0.4952 

# just with predictors correlated to the response greater than 0.3
lake_color <- lake_color %>% select(X_highcor)
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)

```



### vi. 3D plot of PCAs


```{r}
library(scatterplot3d)
lake_color<- pc.scores %>% select(col_hex, paste0("Comp.",c(2,4,7)))

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/PCA/",Sys.Date(),"_PC4_PC2_PC7.png"), height=5, width=7,units = "in", res=500)
scatterplot3d(lake_color$Comp.4, lake_color$Comp.2, lake_color$Comp.7, color=lake_color$col_hex, pch = 16, main = "", xlab = "PC4", ylab = "PC2", zlab = "PC7")
dev.off()

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/PCA/",Sys.Date(),"_PC2_PC7_PC4.png"), height=5, width=7,units = "in", res=500)
scatterplot3d(lake_color$Comp.2, lake_color$Comp.7, lake_color$Comp.4, color =lake_color$col_hex, pch = 16, main = "", xlab = "PC2", ylab = "PC7", zlab = "PC4")
dev.off()

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/PCA/",Sys.Date(),"_PC7_PC4_PC2.png"), height=5, width=7,units = "in", res=500)
scatterplot3d(lake_color$Comp.7, lake_color$Comp.4, lake_color$Comp.2, color =lake_color$col_hex, pch = 16, main = "", xlab = "PC7", ylab = "PC4", zlab = "PC2")
dev.off()




library(scatterplot3d)
lake_color<- pc.scores %>% select(col_hex, paste0("Comp.",c(3,4,5)))

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/PCA/",Sys.Date(),"_PC3_PC4_PC5.png"), height=5, width=7,units = "in", res=500)
scatterplot3d(lake_color$Comp.3, lake_color$Comp.4, lake_color$Comp.5, color=lake_color$col_hex, pch = 16, main = "", xlab = "PC3", ylab = "PC4", zlab = "PC5")
dev.off()

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/PCA/",Sys.Date(),"_PC4_PC5_PC3.png"), height=5, width=7,units = "in", res=500)
scatterplot3d(lake_color$Comp.4, lake_color$Comp.5, lake_color$Comp.3, color=lake_color$col_hex, pch = 16, main = "", xlab = "PC4", ylab = "PC5", zlab = "PC3")
dev.off()


png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/PCA/",Sys.Date(),"_PC5_PC3_PC4.png"), height=5, width=7,units = "in", res=500)
scatterplot3d(lake_color$Comp.5, lake_color$Comp.3, lake_color$Comp.4, color=lake_color$col_hex, pch = 16, main = "", xlab = "PC5", ylab = "PC3", zlab = "PC4")
dev.off()



```



## c. NMDS
```{r}
require(vegan)
require(cluster)
require(phyloseq)
daisy.mat <- as.matrix(daisy(scale(X), metric="gower"))
nmds<-metaMDS(X,distance = "bray",)

```

# distribution of data


pairs.panel.sh
```{bash}
#!/bin/bash
#SBATCH --job-name pairs_panel
#SBATCH --mem=100GB
#SBATCH --time=1-00:00:00
#SBATCH --cpus-per-task=1
#SBATCH --account=microbiome
#SBATCH --output=pairs_panel_%A.out

cd /project/lakecolor/data_analysis/X_Y_distributions
module load arcc/1.0  gcc/12.2.0 
module load r/4.2.2

srun Rscript make_pairs_panels.R
date
```


make_pairs_panels.R
```{r}
load("../../data/2024-03-07_Trends_Prism_Landcover.Rdata")

require(psych)
require(tidyverse)

X_N <- rownames_to_column(X, var = "rownames")
Y_N <- rownames_to_column(Y, var = "rownames")
lake_color <- inner_join(X_N,Y_N,by="rownames") %>% 
        select(-c(trend_cat,trend_cat_switch,lm_slope)) %>%
               column_to_rownames(var = "rownames") %>%
        select(avg_dwl_site, everything()) 

jpeg(paste0(Sys.Date(),"_PairsPanel.pdf"), height=20,width=20,units="in", res=500)
pairs.panels(lake_color,ellipses = F,digits = 3,hist.col = "lightgreen")
dev.off()

```

Shapiro-Wilk normality
mbox text

```



# cluster
Do this in Beartooth since such a large dataset

```{bash}
salloc --mem=110GB --nodes=1 --cpus-per-task=1 --account=microbiome --time=2:00:00

module load arcc/1.0  gcc/12.2.0 
module load r/4.2.2
R
```

```{r}
load("../../data/2024-03-07_Trends_Prism_Landcover.Rdata")

d<-dist(X, method="euclidean")
n=nrow(X); p=ncol(X)

R=cor(X); 
R<-round(R,4)
write.csv(R,paste0(Sys.date(), "_correlation_predcitors.csv"))

```


# random forest

```{r}
require(randomForest)
require(tidyverse)
devtools::install_github("araastat/reprtree")
require(reprtree)

load("../../data/2024-03-07_Trends_Prism_Landcover.Rdata")


set.seed(675)
X_N <- rownames_to_column(X, var = "rownames")
Y_N <- rownames_to_column(Y, var = "rownames")
lake_color <- inner_join(X_N,Y_N,by="rownames") %>% 
        select(-c(trend_cat,avg_dwl_site,lm_slope)) %>%
               column_to_rownames(var = "rownames") %>%
        select(trend_cat_switch, everything()) %>%
        mutate(trend_cat_switch=as.factor(trend_cat_switch))

fit.rF<-randomForest(trend_cat_switch~.,data=lake_color, mtry=sqrt(22))

pdf("Random_Forest_varimpplot.pdf", height=5,width=5)
varImpPlot(fit.rF)
dev.off()


pdf("Random_Forest_errorvstrees.pdf", height=5,width=5)
plot(fit.rF, lwd=2)
dev.off()

pdf("Random_Forest_trees.pdf", height=10,width=10)
plot.getTree(fit.rF)
dev.off()

save.image(paste0(Sys.Date(),"_RandomForestModel.Rdata"))

```


# neural network

```{r}
load("/../../../2024-03-07_Trends_Prism_Landcover.rdata")

require(nnet)
require(tidyverse)


X_N <- rownames_to_column(X, var = "rownames")
Y_N <- rownames_to_column(Y, var = "rownames")
lake_color <- inner_join(X_N,Y_N,by="rownames") %>% 
        select(-c(trend_cat,avg_dwl_site,lm_slope)) %>%
               column_to_rownames(var = "rownames") %>%
        select(trend_cat_switch, everything()) %>%
        mutate(trend_cat_switch=as.factor(trend_cat_switch))

# do this with a subset of 5,000 observations
train_data<-data.frame()
for (i in 1:5000) {
  sampled_rows <- lake_color[sample(nrow(lake_color), size = 1), ]
  train_data <- rbind(train_data, sampled_rows)
}



test_data<-data.frame()
for (i in 1:1000) {
  sampled_rows <- lake_color[sample(nrow(lake_color), size = 1), ]
  test_data <- rbind(test_data, sampled_rows)
}


# Train the nnet model using the training data
nnet_model <- nnet(trend_cat_switch ~., data = lake_color, size = 2, trace = FALSE, maxit = 200)

# Make predictions on the test data
predictions <- predict(nnet_model, newdata = test_data, type = "class")

# Evaluate the predictions (replace 'trend_cat_switch' with your actual response variable)
confusion_matrix <- table(predictions, test_data$trend_cat_switch)


# Extract counts of incorrect assignments
if (is.matrix(confusion_matrix) && nrow(confusion_matrix) == 2 && ncol(confusion_matrix) == 2) {
  # For binary classification
  incorrect_assignments <- confusion_matrix[1, 2] + confusion_matrix[2, 1]
} else if (is.matrix(confusion_matrix)) {
  # For multiclass classification (sum off-diagonal elements)
  incorrect_assignments <- sum(confusion_matrix) - sum(diag(confusion_matrix))
} else {
  stop("Invalid confusion matrix format.")
}

# Print the counts of incorrect assignments
cat("Number of Incorrect Assignments:", incorrect_assignments, "\n")
```



# 4. Figure 3 
## i. add histogram of dates 

Histogram of dates when crossed green/blue and blue to green time series
```{r}

```




# 4. Seasonal variability in lake color

```{r}
names(temporal)
require(ggplot2)
# Get state borders
states <- map_data("state")

temporal_uniq<-temporal %>%
        select(lake_nhdid, lake_lat_decdeg, lake_lon_decdeg, mean_range_dwl_site, lake_waterarea_ha) %>%
        distinct()

map_trends<-ggplot(data = temporal_uniq, aes(x = lake_lon_decdeg, y = lake_lat_decdeg)) +
  geom_point(size = 0.7) +
  scale_color_continuous(values = levels(temporal_uniq$trend_color), 
                     labels = levels(temporal_uniq$trend_cat)) + 
        guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_void() +
  coord_fixed(1.3) +
labs(color = "Lake color trend") +
  geom_polygon(data = states, aes(x = long, y = lat, group = group), fill = "transparent", color = "black", linewidth = 0.5) +
        theme(legend.position = "right",  # Change the legend position to the right
        legend.box.margin = margin(l = -20))


jpeg(paste0("Figures/",Sys.Date(),"_LakeColorTrends_EcoRegionsMap.jpg"),height=6,width=9, res=500, units="in")
map_trends
dev.off()

```







Question - what types of lakes were removed with the temporal dataset?

