---
title: "02_DataAnalysis"
author: "Jordan Von Eggers"
date: "`r Sys.Date()`"
output: pdf_document
---

# 0. Set up on Beartooth 
Request resources, move to directory, and load modules for R
```{bash}
ssh jvonegge@beartooth.arcc.uwyo.edu
cd /project/lakecolor/data_analysis/lakecolor_climate_landcover_alltrends

salloc --mem=110GB --nodes=1 --cpus-per-task=8 --account=lakecolor --time=3:00:00

module load arcc/1.0  gcc/12.2.0 
module load r/4.4.0
R
```

```{r}
require(tidyverse)
#color for manually created legend
bg.fui = tibble(
  ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
  ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
  color = c(
  "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
  "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
  "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04")
)
```





###############################
LINEAR MODELS GOES HERE (USE THE R SCRIPT) 
###############################

- minus 1992-1995
- update the starting and ending file locations


2024-08-02_TEMPORAL_Avg_LMs.R
```{r}
require(tidyverse)
limno<-readRDS("/project/lakecolor/data/01_Compile_Filter_Data/2024-07-30_Limnosat_Lagos.rds")

bg.fui = tibble(
        ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
        ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
        color = c(
                "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
                "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
                "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04")
)

bg.fui<-as.data.frame(bg.fui)

temporal_avg_lms <- limno %>%
        filter(!year%in%c(1992,1993,1994,1995))%>% # remove years 1992-1995 for due to the Mt Pinatubo eruption 
        group_by(lake_nhdid) %>%
        mutate(range_dwl_site=max(dWL)-min(dWL)) %>% 
        ungroup() %>%
        group_by(lake_nhdid,year) %>%
        mutate(avg_dwl_year=mean(dWL), 
               range_dwl_site_year= max(dWL)- min(dWL),
               obs_year=n(),
               obs_days_year=as.numeric(max(as.Date(date)) - min(as.Date(date)))) %>% ungroup() %>%
        group_by(lake_nhdid) %>%
        mutate(mean_range_dwl_site=mean(range_dwl_site_year)) %>% ungroup() %>%
        select(-date,-dWL,-brightness,-normalized_green,-Nir,-lake_year) %>% 
        distinct() %>%
        
        #linear models
        group_by(lake_nhdid) %>%
        mutate(lm_yint=summary(lm(avg_dwl_year~year))$coefficients[1,1],
               lm_pval=summary(lm(avg_dwl_year~year))$coefficients[2,4],
               lm_rsq=summary(lm(avg_dwl_year~year))$r.squared,
               lm_slope=summary(lm(avg_dwl_year~year))$coefficients[2,1],
               residuals=abs(lm(avg_dwl_year~year)$residuals),
               resid_yint=summary(lm(residuals~year))$coefficients[1,1], 
               resid_pval=summary(lm(residuals~year))$coefficients[2,4],
               resid_rsq=summary(lm(residuals~year))$r.squared,
               resid_slope=summary(lm(residuals~year))$coefficients[2,1]) %>% ungroup() %>%
        
        #adding in hex color
        mutate( col_hex = case_when(avg_dwl_year >=bg.fui[1,1] & avg_dwl_year <bg.fui[1,2] ~ bg.fui[1,3],
                                    avg_dwl_year >=bg.fui[2,1] & avg_dwl_year <bg.fui[2,2] ~bg.fui[2,3], 
                                    avg_dwl_year >=bg.fui[3,1] & avg_dwl_year <bg.fui[3,2] ~bg.fui[3,3]   , 
                                    avg_dwl_year >=bg.fui[4,1] & avg_dwl_year <bg.fui[4,2] ~bg.fui[4,3]   , 
                                    avg_dwl_year >=bg.fui[5,1] & avg_dwl_year <bg.fui[5,2] ~bg.fui[5,3]   , 
                                    avg_dwl_year >=bg.fui[6,1] & avg_dwl_year <bg.fui[6,2] ~bg.fui[6,3]   , 
                                    avg_dwl_year >=bg.fui[7,1] & avg_dwl_year <bg.fui[7,2] ~bg.fui[7,3]   , 
                                    avg_dwl_year >=bg.fui[8,1] & avg_dwl_year <bg.fui[8,2] ~bg.fui[8,3]   , 
                                    avg_dwl_year >=bg.fui[9,1] & avg_dwl_year <bg.fui[9,2] ~bg.fui[9,3]   , 
                                    avg_dwl_year >=bg.fui[10,1] & avg_dwl_year <bg.fui[10,2] ~bg.fui[10,3]   , 
                                    avg_dwl_year >=bg.fui[11,1] & avg_dwl_year <bg.fui[11,2] ~bg.fui[11,3]   , 
                                    avg_dwl_year >=bg.fui[12,1] & avg_dwl_year <bg.fui[12,2] ~bg.fui[12,3]   , 
                                    avg_dwl_year >=bg.fui[13,1] & avg_dwl_year <bg.fui[13,2] ~bg.fui[13,3]   , 
                                    avg_dwl_year >=bg.fui[14,1] & avg_dwl_year <bg.fui[14,2] ~bg.fui[14,3]   , 
                                    avg_dwl_year >=bg.fui[15,1] & avg_dwl_year <bg.fui[15,2] ~bg.fui[15,3]   , 
                                    avg_dwl_year >=bg.fui[16,1] & avg_dwl_year <bg.fui[16,2] ~bg.fui[16,3]   , 
                                    avg_dwl_year >=bg.fui[17,1] & avg_dwl_year <bg.fui[17,2] ~bg.fui[17,3]   , 
                                    avg_dwl_year >=bg.fui[18,1] & avg_dwl_year <bg.fui[18,2] ~bg.fui[18,3]   , 
                                    avg_dwl_year >=bg.fui[19,1] & avg_dwl_year <bg.fui[19,2] ~bg.fui[19,3]   , 
                                    avg_dwl_year >=bg.fui[20,1] & avg_dwl_year <bg.fui[20,2] ~bg.fui[20,3]   , 
                                    avg_dwl_year >=bg.fui[21,1] & avg_dwl_year <=bg.fui[21,2] ~bg.fui[21,3] , T~"black")) %>%
        
        #adding in category
        group_by(lake_nhdid) %>%
        mutate(dwl_prior2002=mean(avg_dwl_year[year <= 2002])) %>%
        ungroup() %>%
        
        mutate(trend_cat= case_when(
                lm_pval <= 0.05 & lm_slope >= 0 & dwl_prior2002 < 530 ~ 'Blue -> Greener',
                lm_pval <= 0.05 & lm_slope >= 0 & dwl_prior2002 > 530 ~ 'Intensifying Green/brown',
                lm_pval <= 0.05 & lm_slope <= 0 & dwl_prior2002 > 530 ~ 'Green -> Bluer',
                lm_pval <= 0.05 & lm_slope <= 0 & dwl_prior2002 < 530 ~ 'Intensifying Blue',
                lm_pval > 0.05 & dwl_prior2002 < 530 ~ 'No trend - Blue',
                lm_pval > 0.05 & dwl_prior2002 > 530 ~ 'No trend - Green/brown'
        ),
        trend_cat = factor(
                trend_cat,
                levels = c(
                        'No trend - Blue',
                        'No trend - Green/brown',
                        'Intensifying Blue',
                        'Green -> Bluer',
                        'Intensifying Green/brown',
                        'Blue -> Greener'
                )
        )
        )


write.csv(temporal_avg_lms,paste0(Sys.Date(),"lake_color_linear_models.csv"))
```



run_2024-08-02_TEMPORAL_Avg_LM.sh
```{bash}
#!/bin/bash
#SBATCH --job-name temporal_lms
#SBATCH --mem=100GB
#SBATCH --time=02:00:00
#SBATCH --cpus-per-task=1
#SBATCH --account=microbiome
#SBATCH --output=temporal_lms_%A.out

cd /project/lakecolor/data/02_Data_Analysis
module load arcc/1.0  gcc/12.2.0 
module load r/4.4.0

srun Rscript 2024-08-02_TEMPORAL_Avg_LMs.R
date
```


# LEFT OFF HERE

# 3. compile lake color trends, climate, and landcover
- Add back in ecoregion and any other categorical predictors removed for PCA
- Add in linear model slope of temperature and precipitation over time (done)
- Try without switching categories
- Try with lake colors themselves
- Try with change in land cover environmental dissimilarity (its pairwise comparison), so could select by lake and minimum dissimilarity minus the maximum dissimilarity. Or just take maximum dissimilarity
- Add in development % as low medium high


prism_climate_trends.R
```{r}
require(tidyverse)

# 1. read in climate data from PRISM
prism<-readRDS("../../data/prism_lakepoint_data.rds")
range(prism$year) # range of PRSIM data
#[1] "1984" "2022"

#remove two lakes that have NAs that messes the lm up
prism<-prism[complete.cases(prism),]

# average the mean temperature, the temperature range, rainfall, fo the 39 years of observation
prism_trends <- prism %>%
        arrange(lake_nhdid, year)%>%
        group_by(lake_nhdid) %>%
        mutate(
                # climate trends
                lm_slope_temp=summary(lm(tmean~year))$coefficients[2,1],
                lm_pval_temp=summary(lm(tmean~year))$coefficients[2,4],
                # precipitation
                lm_slope_ppt=summary(lm(ppt~year))$coefficients[2,1],
                lm_pval_ppt=summary(lm(ppt~year))$coefficients[2,4],
                # climate/precip means and ranges
                avg_temp_degC = mean(tmean),
                avg_annual_range_temp_degC = mean(tmax - tmin), # this is tmax and min within a year
                range_ppt = max(ppt)-min(ppt),
                avg_ppt = mean(ppt),
                obs_year_prism = n()) %>%
        ungroup() %>%
        select(lake_nhdid,avg_temp_degC,avg_annual_range_temp_degC, range_ppt, avg_ppt,obs_year_prism, lm_slope_temp, lm_slope_ppt) %>%
        distinct()
dim(prism_trends) # 145083 sites
table(prism_trends$obs_year_prism) # all have 39 years
summary(prism_trends) 


write.csv(prism_trends,"2024-04-25_Prism_climate_trends.csv")
```


```{r}
require(tidyverse)
# read in climate data
prism<-read.csv("2024-04-25_Prism_climate_trends.csv", header=T,row.names=1)

# read in temporal trend data
trends<-read.csv("../../data/2023-12-11_TEMPORAL_Avg_LMs.csv", header=T, row.names=1)
names(trends)
trends$X<-NULL

# rename columns to specify that this is lake color trends we are looking at

trends<-trends %>%
        rename(lm_yint_color=lm_yint,
               lm_pval_color=lm_pval,
               lm_rsq_color=lm_rsq,
               lm_slope_color=lm_slope,
               residuals_color=residuals,
               resid_yint_color=resid_yint,
               resid_pval_color=resid_pval,
               resid_rsq_color=resid_rsq,
               resid_slope_color=resid_slope)

# label lakes that ACTUALLY shifted from blue to green or green to blue 
trend_change <- trends %>%
        group_by(lake_nhdid) %>%
        mutate(dwl_after2002 = mean(avg_dwl_year[year > 2002])) %>%
        ungroup() |>
        mutate(
                trend_cat_switch = ifelse(
                        trend_cat %in% c('Blue -> Greener', 'Green -> Bluer') &
                                dwl_prior2002 > 530 &
                                dwl_after2002 < 530,
                        'Shift to blue',
                        ifelse(
                                trend_cat %in% c('Blue -> Greener', 'Green -> Bluer') &
                                        dwl_prior2002 < 530 &
                                        dwl_after2002 > 530,
                                'Shift to green',
                                trend_cat)))

trend_change %>%
        filter(trend_cat_switch %in%c('Shift to green','Shift to blue'))%>%
        select(lake_nhdid) %>%
        distinct() %>%
        nrow()
# [1] 5133 - same as what Linnea found in code for Figure 3


trends_avg <- trend_change %>%
        group_by(lake_nhdid) %>%
        mutate(avg_obs_site = mean(obs_year),
               avg_obs_days_site = mean(obs_days_year),
               avg_dwl_site = mean(avg_dwl_year),
               avg_range_dwl_site = mean(range_dwl_site_year)) %>%
                       ungroup()%>%
        select(
                lake_nhdid,
                epanutr_zoneid,
                hu12_zoneid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                lake_islandperimeter_m,
                lake_connectivity_class,
                lake_connectivity_permanent,
                lake_totalperimeter_m,
                ws_area_ha,
                ws_lake_arearatio,
                range_dwl_site,
                mean_range_dwl_site,
                lm_yint_color,
                lm_pval_color,
                lm_rsq_color,
                lm_slope_color,
                resid_yint_color,
                resid_pval_color,
                resid_rsq_color,
                resid_slope_color,
                dwl_prior2002,
                trend_cat,
                trend_cat_switch,
                avg_obs_site,
                avg_obs_days_site,
                avg_dwl_site,
                avg_range_dwl_site
        )%>%
        distinct()
dim(trends_avg)
#[1] 81007    31

# join the trend data and prism data
trends_prism <- inner_join(trends_avg,prism, by= join_by(lake_nhdid))
dim(trends_prism)
#[1] 81007    38

# read in landcover data, filter by lakes we have in trends, and average all columns
landcover <- readRDS("/project/lakecolor/data/archive/2023-09-29_lagos_landcover_data.rds")

landcover_avg_diff <- landcover %>%
        arrange(lake_nhdid, year) %>%
                select(lake_nhdid,
                nlcd_barren31_pct,
                nlcd_cultcrop82_pct,
                nlcd_devhi24_pct,
                nlcd_devlow22_pct,
                nlcd_devmed23_pct,
                nlcd_devopen21_pct,
                nlcd_forcon42_pct,
                nlcd_fordec41_pct,
                nlcd_formix43_pct,
                nlcd_grass71_pct,
                nlcd_icesnow12_pct,
                nlcd_openwater11_pct,
                nlcd_past81_pct,
                nlcd_shrub52_pct,
                nlcd_wetemerg95_pct,
                nlcd_wetwood90_pct
        ) %>%         group_by(lake_nhdid) %>%
        mutate(
#add in averaged column
        across(starts_with("nlcd"), list(avg = ~ mean(.)), .names = "avg_{.col}"),
nlcd_barren31_diff = nlcd_barren31_pct[7]-nlcd_barren31_pct[1],
nlcd_cultcrop82_diff = nlcd_cultcrop82_pct[7]-nlcd_cultcrop82_pct[1],
nlcd_devhi24_diff = nlcd_devhi24_pct[7]-nlcd_devhi24_pct[1],
nlcd_devlow22_diff = nlcd_devlow22_pct[7]-nlcd_devlow22_pct[1],
nlcd_devmed23_diff = nlcd_devmed23_pct[7]-nlcd_devmed23_pct[1],
nlcd_devopen21_diff = nlcd_devopen21_pct[7]-nlcd_devopen21_pct[1],
nlcd_forcon42_diff = nlcd_forcon42_pct[7]-nlcd_forcon42_pct[1],
nlcd_fordec41_diff = nlcd_fordec41_pct[7]-nlcd_fordec41_pct[1],
nlcd_formix43_diff = nlcd_formix43_pct[7]-nlcd_formix43_pct[1],
nlcd_grass71_diff = nlcd_grass71_pct[7]-nlcd_grass71_pct[1],
nlcd_icesnow12_diff = nlcd_icesnow12_pct[7]-nlcd_icesnow12_pct[1],
nlcd_openwater11_diff = nlcd_openwater11_pct[7]-nlcd_openwater11_pct[1],
nlcd_past81_diff = nlcd_past81_pct[7]-nlcd_past81_pct[1],
nlcd_shrub52_diff = nlcd_shrub52_pct[7]-nlcd_shrub52_pct[1],
nlcd_wetemerg95_diff = nlcd_wetemerg95_pct[7]-nlcd_wetemerg95_pct[1],
nlcd_wetwood90_diff = nlcd_wetwood90_pct[7]-nlcd_wetwood90_pct[1],

nlcd_summed_forest_diff= sum(nlcd_forcon42_pct[7],nlcd_fordec41_pct[7], nlcd_formix43_pct[7])- sum(nlcd_forcon42_pct[1],nlcd_fordec41_pct[1], nlcd_formix43_pct[1]),
nlcd_summed_wetland_diff = sum(nlcd_wetwood90_pct[7],nlcd_wetemerg95_pct[7])- sum(nlcd_wetwood90_pct[1],nlcd_wetemerg95_pct[1]),
nlcd_summed_agr_diff = sum(nlcd_cultcrop82_pct[7],nlcd_past81_pct[7])- sum(nlcd_cultcrop82_pct[1],nlcd_past81_pct[1]),
nlcd_summed_dev_diff= sum(nlcd_devlow22_pct[7],nlcd_devhi24_pct[7],nlcd_devmed23_pct[7], nlcd_devopen21_pct[7])- sum(nlcd_devlow22_pct[1],nlcd_devhi24_pct[1],nlcd_devmed23_pct[1], nlcd_devopen21_pct[1])
                ) %>% ungroup() %>% 
                select(-c(
                nlcd_barren31_pct,
                nlcd_cultcrop82_pct,
                nlcd_devhi24_pct,
                nlcd_devlow22_pct,
                nlcd_devmed23_pct,
                nlcd_devopen21_pct,
                nlcd_forcon42_pct,
                nlcd_fordec41_pct,
                nlcd_formix43_pct,
                nlcd_grass71_pct,
                nlcd_icesnow12_pct,
                nlcd_openwater11_pct,
                nlcd_past81_pct,
                nlcd_shrub52_pct,
                nlcd_wetemerg95_pct,
                nlcd_wetwood90_pct
                )
        ) %>% # get rid of year and data from individual years
        distinct() %>%
        rowwise() |>
  mutate(avg_nlcd_dev_pct = sum(
    avg_nlcd_devlow22_pct,
    avg_nlcd_devmed23_pct,
    avg_nlcd_devhi24_pct,
    avg_nlcd_devopen21_pct
  ),
  avg_nlcd_ag_pct = sum(
    avg_nlcd_cultcrop82_pct,
    avg_nlcd_past81_pct
  ),
  avg_nlcd_forest_pct = sum(
    avg_nlcd_forcon42_pct,
    avg_nlcd_fordec41_pct,
    avg_nlcd_formix43_pct
  ),
  avg_nlcd_wetland_pct = sum(
    avg_nlcd_wetwood90_pct,
    avg_nlcd_wetemerg95_pct
  )) |>
  ungroup()


#merge landcover with prism and lake color trends
trends_prism_landcover <- inner_join(trends_prism,landcover_avg_diff, by= join_by(lake_nhdid))
dim(trends_prism_landcover)
#[1] 81006    78

write.csv(trends_prism_landcover,paste0(Sys.Date(),"_lakecolor_climate_landcover_trends.csv"))
```

