---
title: "Figures_4-6"
author: "Jordan Von Eggers"
date: "`r Sys.Date()`"
output: html_document
---

```{bash}
ssh jvonegge@beartooth.arcc.uwyo.edu
cd /project/lakecolor/data/

salloc --mem=110GB --nodes=1 --cpus-per-task=8 --account=lakecolor --time=3:00:00

module load arcc/1.0  gcc/12.2.0 
module load r/4.4.0
R
```

# Custom theme
```{r}
  custom_theme <- function() {
  theme_bw() +
    theme(
      text = element_text(color = "black", size = 13),
      axis.text = element_text(color = "black",size = 13),
      axis.title = element_text(color = "black",size = 13),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      legend.text = element_text(size = 13),
      legend.title = element_text(size = 13),
      strip.text = element_text(size = 13, color = "black"))}
```


# NEW
# add in connectivity class and other data
```{r}
# finalized dataset with only lake nhd id
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-08-29_lakecolor_climate_landcover_trends.rds") 

# new data we want to include from Lagos (https://portal.edirepository.org/nis/metadataviewer?packageid=edi.854.1)
lagos_lake_char<-read.csv("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/lake_characteristics.csv", header=T)

# dataframe with lagos lake ids and lake nhd ids
connect<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-07-30_Limnosat_Lagos.rds") %>% select(lake_nhdid,lagoslakeid) %>% distinct(.)

lagos_lake_char_connect<-left_join(connect, lagos_lake_char, join_by(lagoslakeid)) %>% select(lagoslakeid, lake_nhdid, lake_lakes4ha_upstream_ha,lake_lakes4ha_upstream_n,lake_lakes1ha_upstream_ha,lake_lakes1ha_upstream_n,lake_lakes10ha_upstream_n,lake_lakes10ha_upstream_ha,lake_glaciatedlatewisc)

lk<-left_join(lk,lagos_lake_char_connect, join_by(lake_nhdid))

write_rds(lk,"/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-11-11_lakecolor_climate_landcover_trends_update.rds")
rm(list=ls())
```
Description of new added variables from https://lagoslakes.wordpress.com/wp-content/uploads/2021/12/locus_user_guide.pdf
"Connectivity variables reflect freshwater connections between each lake and any
inflowing streams and upstream lakes > 10 ha in surface area based on lake and stream
connections in the NHDPlus HR Beta (§ 3.5.2). Connectivity class reflects the strength of each
lake’s surface connectivity with other lakes and with streams. The six connectivity classes are
distinguished by three features: the presence of inflows, presence of outflows and connections to
upstream lakes > 10 ha in area. ‘Isolated’ lakes lack both inflows and outflows; ‘Headwater’
lakes have an outflow but no inflow; ‘Drainage’ lakes have both inflow(s) and outflow(s) and
lack upstream connected lake > 10 ha in area; ‘DrainageLk’ lakes have inflow(s) and outflow(s)
and have upstream connected lake(s) > 10 ha; ‘Terminal’ lakes have inflows but no outflows and
lack upstream connected lake > 10 ha in area; and ‘TerminalLk’ lakes have inflow(s) but no
outflow(s) and do have upstream connected lake(s) > 10 ha. Because the stream network labeled
each stream segment as permanent or intermittent, we calculated two variables for connectivity
class: (1) using all available stream connections in the network and (2) using those defined only
by permanent flow. To highlight lakes with connectivity class types that differ depending on flow
permanence, we created a variable that indicates whether a connectivity class for a given lake
fluctuates with flow permanence. A second set of freshwater connectivity metrics was quantified
as the number and total area of connected upstream lakes > 1, 4, and 10 ha. There are no
observation-level flags for freshwater connectivity.
Finally, we include a variable that indicates whether or not any part of the focal lake was
glaciated during the Late Wisconsin glaciation period. There are no observation-level flags for
glaciation status."


# MS stats
```{r}
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-08-29_lakecolor_climate_landcover_trends.rds") 

# how many lakes are small
nrow(lk[lk$lake_waterarea_ha<10,])/nrow(lk)
# 0.5281433 ot 52.8% of lakes are small!

lk_yrs<-read.csv("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-08-29_lake_color_linear_models.csv")
# what is the range and average duration of the study period
year_summary<-lk_yrs %>% group_by(lake_nhdid) %>%
        summarize(yr_range=max(year)-min(year),
                  average_duration=n())

table(year_summary$yr_range)
37+4

#   25    29    30    31    32    33    34    35    36    37 
#  171    27   337   489   597  1075  2994 11416 11818 52926 

25+4


mean(year_summary$average_duration)

as.data.frame(table(lk$trend_cat)/nrow(lk))

# Blue -> Greener	0.05455167			
# Green -> Bluer	0.18487703			
# Intensifying Blue	0.10157729			
# Intensifying Green/brown	0.07098437			
# No trend - Blue	0.25377219			
# No trend - Green/brown	0.33423744

lk <- lk %>% 
  mutate(color_category = case_when(
    avg_dwl_site > 530 ~ "Green", 
    avg_dwl_site < 530 ~ "Blue"))
table(lk$color_category)/nrow(lk)
#      Blue     Green 
# 0.4278855 0.5721145 
```



# Figure 4 
Random forest with training dataset
```{r}
require(randomForest)
require(tidyverse)
require(caret)
library(paletteer)
require(ggplot2)
```

## Green vs blue random forest
```{r}
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-11-11_lakecolor_climate_landcover_trends_update.rds") 
nrow(lk) # 81849

# looking at blue vs green lakes
lake_color <- lk %>% 
column_to_rownames(var = "lake_nhdid") %>% 
  mutate(color_category = case_when(
    avg_dwl_site > 530 ~ "Green", 
    avg_dwl_site < 530 ~ "Blue"),
  size_category= case_when(
        lake_waterarea_ha <10 ~"small",
        lake_waterarea_ha>=10~"large")) %>%
        # convert to factors
  mutate(color_category = as.factor(color_category),
         size_category=as.factor(size_category),
         lake_connectivity_class=as.factor(lake_connectivity_class),
         epanutr_zoneid = as.factor(epanutr_zoneid),
         lake_glaciatedlatewisc = as.factor(lake_glaciatedlatewisc)) %>% 
select(        # random forest category
                color_category,
                # lake size and location variables
                epanutr_zoneid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                ws_area_ha,
                lake_glaciatedlatewisc,
                size_category,
                # connectivity variables
                lake_connectivity_class,
                lake_lakes1ha_upstream_ha,
                lake_lakes1ha_upstream_n,
                # climate variables
                avg_temp_degC,
                avg_ppt,
                lm_slope_temp,
                lm_slope_ppt,
                avg_annual_range_temp_degC,
                # landcover vars
                nlcd_summed_forest_diff,
                nlcd_summed_wetland_diff,
                nlcd_summed_agr_diff,
                nlcd_summed_dev_diff,
                avg_nlcd_dev_pct,
                avg_nlcd_ag_pct,
                avg_nlcd_forest_pct,
                avg_nlcd_wetland_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_barren31_pct,
                avg_nlcd_openwater11_pct
        ) %>% na.omit() # some NAs for watershed area in the TX and GA lakes

nrow(lake_color)
#[1] 81773


# count number of observations for blue and green by eco region
category_counts <- lake_color %>%
  group_by(epanutr_zoneid, color_category) %>% 
  summarise(n = n(), .groups = 'drop')

# subset equally lakes within each ecoregion
set.seed(111)
train <- lake_color %>%
  group_by(epanutr_zoneid) %>%
  sample_frac(.8) %>% #80% of data will be used for training
  ungroup() 
 
validate <- lake_color %>%
   anti_join(train)

# First rF looking at the color category
set.seed(111)
(rF_color <- randomForest(color_category ~ ., data = train, mtry = 6))
```
Call:
 randomForest(formula = color_category ~ ., data = train, mtry = 6) 
               Type of random forest: classification
                     Number of trees: 500
No. of variables tried at each split: 6

        OOB estimate of  error rate: 14.78%
Confusion matrix:
       Blue Green class.error
Blue  23041  4975   0.1775771
Green  4691 32711   0.1254211

```{r}
rF_color$err.rate[nrow(rF_color$err.rate), ]

# percent of correct classifications
round((1- rF_color$err.rate[nrow(rF_color$err.rate), 2])*100,digits=1) 
round((1- rF_color$err.rate[nrow(rF_color$err.rate), 3])*100,digits=1) 

rF_color_pred <- predict(rF_color, validate)
rF_color_pred_confus <-table(observed = validate$color_category, predicted = rF_color_pred)

rm(lake_color);rm(lk)

```
 OOB      Blue     Green 
0.1477575 0.1775771 0.1254211 
Blue 
82.2 
Green 
 87.5 

```{r}
cm <- confusionMatrix(rF_color_pred,validate$color_category, dnn=c('Predicted', 'Observed'))

accuracy <- cm$overall

cm <- as.data.frame(cm$table) |>
  group_by(Observed) |>
  mutate(prop = Freq/sum(Freq),
         prop = round(prop,2))

custom_theme <- function() {
  theme_bw() +
    theme(
      text = element_text(color = "black", size = 13),
      axis.text = element_text(color = "black",size = 13),
      axis.title = element_text(color = "black",size = 13),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      legend.text = element_text(size = 13),
      legend.title = element_text(size = 13),
      strip.text = element_text(size = 13, color = "black"))}
  
ggplot(cm, aes(x = Observed, y = Predicted, fill=Freq)) +
  geom_tile(color="black") +
  scale_x_discrete(expand = c(0, 0))+ #remove white space
  scale_y_discrete(expand = c(0, 0))+ #remove white space
  scale_fill_gradient(low="white", high="orangered",
                      name="Frequency") +
  geom_text(aes(label = paste0("n=",Freq)), vjust = .5,  alpha = 1, size=5) +
  geom_text(aes(label = paste0(prop*100,"%")), vjust = 2.0,  alpha = 1, size=4) +
  custom_theme() +
  theme(legend.position = 'none') +
  labs(title = paste0("Accuracy: ",round(accuracy,2))) 

ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/CM_rF_color.png",width=3.7, height=2.5, units="in",dpi = 300)


# create prettier lollipop plot for the color 
rF_imp<-as.data.frame(importance(rF_color)) %>%
       rownames_to_column(.,var="variables")

# update names, fist print as the string that is easy to edit
rF_imp$variables

rF_imp$newvarnames<-c("Ecoregion", "Lake elevation (m)", "Latitude (DD)", "Longitude (DD)", "Lake area (ha)", "Lake perimeter (m)", "Watershed area (ha)","Glaciated", "Lake size (small/large)",  "Connectivity class"  , "Upstream lakes <= 1 ha (ha)" , "Upstream lakes <= 1 ha (no.)"  ,"Average annual air\ntemperature (deg C)", "Average precipitation (ppt)", "Slope of temperature over time", "Slope of precipitation over time", "Average annual air temperature range (deg C)", "Difference in forest 2016-2001 (%)", "Difference in wetland 2016-2001 (%)", "Difference in agriculture 2016-2001 (%)", "Difference in development 2016-2001 (%)", "Average development (%)", "Average agriculture (%)", "Average forest (%)", "Average wetland (%)", "Average ice and snow (%)", "Average shrub (%)", "Average barren (%)", "Average open water (%)")

rF_imp <- rF_imp %>% arrange(MeanDecreaseGini) %>%  mutate(newvarnames = factor(newvarnames, levels = newvarnames)) 
write.csv(rF_imp, "/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/Figure4_rF_importance_color.csv")

rF_imp <- rF_imp %>% slice_tail(n=5) # just plot the top 5


ggplot(rF_imp, aes(y = MeanDecreaseGini, x = newvarnames)) +
  geom_segment(aes(x = newvarnames, xend = newvarnames, y = 0, yend = MeanDecreaseGini), color = "black") +
  geom_point(aes(color = MeanDecreaseGini), size = 4) +
  scale_color_paletteer_c("grDevices::Zissou 1", n = 30, direction = 1) +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
            panel.grid.minor.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(), 
    axis.text = element_text(color = "black", size = 15),  
    axis.title = element_text(color = "black", size = 15), 
    legend.position = "none",
    plot.margin = unit(c(0.2, 0.6, 0.2, 0.2), "cm")) + 
  xlab("") + 
  ylab("Mean decrease in Gini")


ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/Fig4_Importance_rF_color.png",width=6.7, height=3, units="in",dpi = 300)
nrow(train)+nrow(validate)
```
total number of lakes:
[1] 81773


## Trending green, trending blue, no trend random forest
```{r}
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-11-11_lakecolor_climate_landcover_trends_update.rds") 
nrow(lk) 
#81849

# looking at basic trend categories (trending blue, trending green, not trending)
lake_color <- lk %>% 
column_to_rownames(var = "lake_nhdid") %>% 
mutate(basic_trend_cat=case_when(lm_slope_color > 0.4 & lm_pval_color <= 0.05~ "Trending green", 
                              lm_slope_color < (-0.4) & lm_pval_color <= 0.05~ "Trending blue", 
                              lm_pval_color > 0.05 ~ "No trend",
                              lm_slope_color < 0.4 &  lm_slope_color > (-0.4) & lm_pval_color <= 0.05~ NA_character_),
       size_category= case_when(
        lake_waterarea_ha <10 ~"small",
        lake_waterarea_ha>=10~"large")) %>%
        mutate(basic_trend_cat=as.factor(basic_trend_cat),
               size_category=as.factor(size_category),
               lake_connectivity_class=as.factor(lake_connectivity_class),
               epanutr_zoneid = as.factor(epanutr_zoneid),
               lake_glaciatedlatewisc = as.factor(lake_glaciatedlatewisc)) %>% 
        select( # random forest category
                basic_trend_cat,
                # lake size and location variables
                epanutr_zoneid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                ws_area_ha,
                lake_glaciatedlatewisc,
                size_category,
                # connectivity variables
                lake_connectivity_class,
                lake_lakes1ha_upstream_ha,
                lake_lakes1ha_upstream_n,
                # climate variables
                avg_temp_degC,
                avg_ppt,
                lm_slope_temp,
                lm_slope_ppt,
                avg_annual_range_temp_degC,
                # landcover vars
                nlcd_summed_forest_diff,
                nlcd_summed_wetland_diff,
                nlcd_summed_agr_diff,
                nlcd_summed_dev_diff,
                avg_nlcd_dev_pct,
                avg_nlcd_ag_pct,
                avg_nlcd_forest_pct,
                avg_nlcd_wetland_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_barren31_pct,
                avg_nlcd_openwater11_pct
        ) %>% na.omit() # some NAs for watershed area in the TX and GA lakes



nrow(lake_color)
#[1] 60723

# count number of observations for blue and green and no trend by eco region
category_counts <- lake_color %>%
  group_by(epanutr_zoneid, basic_trend_cat) %>% 
  summarise(n = n(), .groups = 'drop')

# what is the minimum number of observations of one of the categories to subset by
print("min(category_counts$n)")
min(category_counts$n)


# sample XX lakes from each ecoregion to make the number of trend categories even
data_subset <- lake_color %>%
  group_by(epanutr_zoneid, basic_trend_cat) %>%
  slice_sample(n = min(category_counts$n), replace = FALSE)  


set.seed(111)
train <- data_subset %>%
  group_by(epanutr_zoneid) %>%
  sample_frac(.8) %>% #80% of data will be used for training
  ungroup() 
 
validate <- data_subset %>%
   anti_join(train)

set.seed(111)
(rF_trend<-randomForest(basic_trend_cat~.,data=train, mtry=6))
```
[1] 81849
[1] 66006
[1] "min(category_counts$n)"
[1] 148

Call:
 randomForest(formula = basic_trend_cat ~ ., data = train, mtry = 6) 
               Type of random forest: classification
                     Number of trees: 500
No. of variables tried at each split: 6

        OOB estimate of  error rate: 52.86%
Confusion matrix:
               No trend Trending blue Trending green class.error
No trend            431           323            322   0.5994424
Trending blue       291           558            197   0.4665392
Trending green      315           241            517   0.5181733



```{r}
rF_trend$err.rate[nrow(rF_trend$err.rate), ]

# percent of correct classifications
round((1- rF_trend$err.rate[nrow(rF_trend$err.rate), 2])*100,digits=1) 
round((1- rF_trend$err.rate[nrow(rF_trend$err.rate), 3])*100,digits=1) 
round((1- rF_trend$err.rate[nrow(rF_trend$err.rate), 4])*100,digits=1) 

rm(lake_color);rm(lk)
```
           OOB       No trend  Trending blue Trending green 
     0.5286385      0.5994424      0.4665392      0.5181733 
No trend 
    40.1 
Trending blue 
         53.3 
Trending green 
          48.2 



```{r}
rF_trend_pred <- predict(rF_trend, validate)
rF_trend_pred_confus <-table(observed =  validate$basic_trend_cat, predicted = rF_trend_pred)

cm <- confusionMatrix(rF_trend_pred,validate$basic_trend_cat, dnn=c('Predicted', 'Observed'))

accuracy <- cm$overall

cm <- as.data.frame(cm$table) |>
  group_by(Observed) |>
  mutate(prop = Freq/sum(Freq),
         prop = round(prop,2))

 custom_theme <- function() {
  theme_bw() +
    theme(
      text = element_text(color = "black", size = 13),
      axis.text = element_text(color = "black",size = 13),
      axis.title = element_text(color = "black",size = 13),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      legend.text = element_text(size = 13),
      legend.title = element_text(size = 13),
      strip.text = element_text(size = 13, color = "black"))}

ggplot(cm, aes(x = Observed, y = Predicted, fill=Freq)) +
  geom_tile(color="black") +
  scale_x_discrete(expand = c(0, 0))+ #remove white space
  scale_y_discrete(expand = c(0, 0))+ #remove white space
  scale_fill_gradient(low="white", high="orangered",
                      name="Frequency") +
  geom_text(aes(label = paste0("n=",Freq)), vjust = .5,  alpha = 1, size=5) +
  geom_text(aes(label = paste0(prop*100,"%")), vjust = 2.0,  alpha = 1, size=4) +
  custom_theme() +
  theme(legend.position = 'none') +
  labs(title = paste0("Accuracy: ",round(accuracy,2))) 

ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/CM_rF_trend.png",width=6, height=2.5, units="in",dpi = 300)



# create prettier lollipop plot 
rF_imp<-as.data.frame(importance(rF_trend)) %>%
       rownames_to_column(.,var="variables")


rF_imp$variables

rF_imp$newvarnames<-c("Ecoregion", "Lake elevation (m)", "Latitude (DD)", "Longitude (DD)", "Lake area (ha)", "Lake perimeter (m)", "Watershed area (ha)","Glaciated", "Lake size (small/large)",  "Connectivity class"  , "Upstream lakes <= 1 ha (ha)" , "Upstream lakes <= 1 ha (no.)"  ,"Average annual air\ntemperature (deg C)", "Average precipitation (ppt)", "Slope of air\ntemperature over time", "Slope of precipitation over time", "Average annual air temperature range (deg C)", "Difference in forest 2016-2001 (%)", "Difference in wetland 2016-2001 (%)", "Difference in agriculture 2016-2001 (%)", "Difference in development 2016-2001 (%)", "Average development (%)", "Average agriculture (%)", "Average forest (%)", "Average wetland (%)", "Average ice and snow (%)", "Average shrub (%)", "Average barren (%)", "Average open water (%)")

rF_imp <- rF_imp %>% arrange(MeanDecreaseGini) %>%  mutate(newvarnames = factor(newvarnames, levels = newvarnames)) 
write.csv(rF_imp, "/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/Figure4_rF_importance_trend.csv")

rF_imp <- rF_imp %>% slice_tail(n=5) # just plot the top 5

ggplot(rF_imp, aes(y = MeanDecreaseGini, x = newvarnames)) +
  geom_segment(aes(x = newvarnames, xend = newvarnames, y = 0, yend = MeanDecreaseGini), color = "black") +
  geom_point(aes(color = MeanDecreaseGini), size = 4) +
scale_color_paletteer_c("grDevices::Zissou 1", n = 30, direction=1) +
        theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(), 
axis.text = element_text(color = "black", size=15),  
    axis.title = element_text(color = "black",size=15),  
     legend.position = "none",
 plot.margin = unit(c(0.2, 0.6, 0.2, 0.2), "cm")) + 
  xlab("") + 
  ylab("Mean decrease in Gini") +
        ylim(c(0,150))


ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/Fig4_Importance_rF_trend.png",width=6.9, height=3, units="in",dpi = 300)

nrow(train)+nrow(validate)
#3996
```


# Figure 5


```{r}
require(patchwork)
custom_theme <- function() {
 theme_classic() +
    theme(
      text = element_text(color = "black", size = 13),
      axis.text = element_text(color = "black",size = 13),
      axis.title = element_text(color = "black",size = 13),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      legend.text = element_text(size = 13),
      legend.title = element_text(size = 13),
      strip.text = element_text(size = 13, color = "black"))}

lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-11-11_lakecolor_climate_landcover_trends_update.rds")
nrow(lk)
#[1] 81849

# looking at blue vs green lakes
lake_color <- lk %>% select(-c(trend_cat, trend_cat_switch, lm_slope_color, lm_pval_color)) %>% 
  mutate(color_category = case_when(
    avg_dwl_site > 530 ~ "Green", 
    avg_dwl_site < 530 ~ "Blue"
  )) %>%
  mutate(color_category = as.factor(color_category)) %>% 
  select(-avg_dwl_site, -mean_range_dwl_site)

# Reshape data to long format
lake_color_long <- lake_color %>%
  pivot_longer(cols = c("avg_nlcd_forest_pct", "lake_lon_decdeg", "lake_lat_decdeg", "avg_temp_degC"), 
               names_to = "variable", 
               values_to = "value")

# Create a named vector for renaming the variables
new_names <- c(
  "avg_nlcd_forest_pct" = "Average forest (%)",
  "lake_lon_decdeg" = "Longitude (DD)",
  "lake_lat_decdeg" = "Latitude (DD)",
  "avg_temp_degC" = "Average annual air\ntemperature (deg C)"
)

# Rename the variables using recode
lake_color_long <- lake_color_long %>%
  mutate(variable = recode(variable, !!!new_names))


summary_stats <- lake_color_long %>%
  group_by(variable, color_category) %>% 
  summarise(
    mean_value = round(mean(value, na.rm = TRUE),digits=2),      
    sd_value = round(sd(value, na.rm = TRUE), digits=2)          
  )


# checking to make sure all t-tests are significant
t_test_results <- lake_color_long %>%
  group_by(variable) %>% # Group by variable
  summarise(
    t_test = list(t.test(value ~ color_category, data = pick(value, color_category))) # Specify columns in pick()
  ) %>%
  mutate(
    t_statistic = sapply(t_test, function(x) x$statistic),   # Extract t-statistic
    p_value = sapply(t_test, function(x) x$p.value),         # Extract p-value
    conf_low = sapply(t_test, function(x) x$conf.int[1]),    # Extract confidence interval lower bound
    conf_high = sapply(t_test, function(x) x$conf.int[2])    # Extract confidence interval upper bound
  ) %>%
  select(-t_test) %>%  # Remove the list column if you just want the summary statistics
        mutate(pval_lessthan0.5=ifelse(p_value<0.05, "Yes","No"))

write.csv(summary_stats, "/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/Figure5_summary_stats_color.csv")

# Create density plots
plt1<-ggplot(lake_color_long, aes(x = value, fill = color_category)) +
  geom_density(alpha = 0.7) +
  facet_wrap(~ variable, nrow=2,scales = "free") +
  labs(x = "",
       y = "Density",
       fill="") +
          scale_fill_manual(values = c("Green" = "#759e72", "Blue"="#327cbb")) +
custom_theme() +
  theme(legend.position = "top") 
        





lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-11-11_lakecolor_climate_landcover_trends_update.rds")
nrow(lk)
#[1] 81849

# looking at blue vs green lakes
lake_color <- lk %>% select(-c(trend_cat, trend_cat_switch,-avg_dwl_site, -mean_range_dwl_site)) %>% 
  mutate(basic_trend_cat=case_when(lm_slope_color > 0.4 & lm_pval_color <= 0.05~ "Trending green", 
                              lm_slope_color < (-0.4) & lm_pval_color <= 0.05~ "Trending blue", 
                              lm_pval_color > 0.05 ~ "No trend",
                              lm_slope_color < 0.4 &  lm_slope_color > (-0.4) & lm_pval_color <= 0.05~ NA_character_
  )) %>%  mutate(basic_trend_cat = as.factor(basic_trend_cat)) %>% filter(is.na(basic_trend_cat)==F)

# Log transform lake area and watershed area
lake_color<- lake_color %>%
        mutate(lake_waterarea_ha=log(lake_waterarea_ha),
               ws_area_ha=log(ws_area_ha),
               lake_perimeter_m=log(lake_perimeter_m))

# Reshape data to long format
lake_color_long <- lake_color %>%
  pivot_longer(cols = c("lm_slope_temp",  "lake_perimeter_m", "ws_area_ha", "lake_elevation_m", "lake_waterarea_ha"), 
               names_to = "variable", 
               values_to = "value")

# Create a named vector for renaming the variables
new_names <- c(
  "lake_perimeter_m" = "log(Lake perimeter (m))",
  "lm_slope_temp" = "Sloe of average air\ntemperature (deg C)",
  "ws_area_ha" = "log(Watershed area (ha))",
  "lake_elevation_m" = "Lake elevation (m)",
  "lake_waterarea_ha" = "log(Lake area (ha))"
)

# Rename the variables using recode
lake_color_long <- lake_color_long %>%
  mutate(variable = recode(variable, !!!new_names))



summary_stats <- lake_color_long %>%
  group_by(variable, basic_trend_cat) %>% 
  summarise(
    mean_value = round(mean(value, na.rm = TRUE),digits=2),      
    sd_value = round(sd(value, na.rm = TRUE), digits=2)          
  )

write.csv(summary_stats, "/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/Figure5_summary_stats_trend.csv")

  custom_theme <- function() {
 theme_classic() +
    theme(
      text = element_text(color = "black", size = 13),
      axis.text = element_text(color = "black",size = 13),
      axis.title = element_text(color = "black",size = 13),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      legend.text = element_text(size = 13),
      legend.title = element_text(size = 13),
      strip.text = element_text(size = 13, color = "black"))}

# Create density plots
plt2<-ggplot(lake_color_long, aes(x = value, fill = basic_trend_cat)) +
  geom_density(alpha = 0.7) +
  facet_wrap(~ variable, nrow=3,scales = "free") +
  labs(x = "",
       y = "Density",
       fill="") +
          scale_fill_manual(values = c("Trending green" = "#759e72", "Trending blue"="#327cbb", "No trend"="gray")) +
custom_theme() +
  theme(legend.position = "top") 
        
(plt1 / plt2)+
  plot_layout( height= c(2,3.3)) + 
  plot_annotation(tag_levels = 'a')


ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure5/Figure5_Density_plots_important_vars.png",width=7, height=11, units="in",dpi = 200)
```





# Figure 6
## PCA
```{r}
require(tidyverse)
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/2024-11-11_lakecolor_climate_landcover_trends_update.rds")
nrow(lk)
#[1] 81849

# select the variables you want to work on, but it will automatically grab date and lake_ndhid
X <- lk %>%
        mutate(size_category= case_when(
                lake_waterarea_ha <10~0,
                lake_waterarea_ha>=10~1),
        glaciated_category=case_when(
                lake_glaciatedlatewisc=="Glaciated"~1,
                lake_glaciatedlatewisc=="Not_Glaciated"~0)) %>%
        select(
                # lake size and location variables
                lake_nhdid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                ws_area_ha,
                glaciated_category,               
                size_category,
                # connectivity variables
                lake_lakes1ha_upstream_ha,
                lake_lakes1ha_upstream_n,
                # climate variables
                avg_temp_degC,
                avg_ppt,
                lm_slope_temp,
                lm_slope_ppt,
                avg_annual_range_temp_degC,
                # landcover vars
                nlcd_summed_forest_diff,
                nlcd_summed_wetland_diff,
                nlcd_summed_agr_diff,
                nlcd_summed_dev_diff,
                avg_nlcd_dev_pct,
                avg_nlcd_ag_pct,
                avg_nlcd_forest_pct,
                avg_nlcd_wetland_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_barren31_pct,
                avg_nlcd_openwater11_pct
        )
dim(X)
#[1] 81849    28

X <- X %>%
       column_to_rownames(var = "lake_nhdid") %>% distinct() %>% drop_na()
dim(X)
#[1] 81773    27

Y <- lk %>%
               filter(lake_nhdid%in%rownames(X))%>%
        select(lake_nhdid,avg_dwl_site, trend_cat, trend_cat_switch, lm_slope_color, lm_pval_color)
 
Y <- Y %>%
       column_to_rownames(var = "lake_nhdid") %>% distinct() %>% drop_na()
dim(Y)
#[1] 81772     5

X <- X %>% filter(rownames(.)%in%rownames(Y))
dim(X)
#[1]  81772    27

rm(list=ls()[! ls() %in% c("Y", "X")])

# run PCA
#R <- cor(X)
#round(R,4)

pc <- princomp(X, cor=T)
```


```{r}
(spc <- summary(pc, loadings=T, cutoff=0.3))
#lambda <- pc$sd^2
plot(pc, type="lines", main = "")
```
Keep 8 PCs

```{r}
for (i in 1:9){
print(paste0("Principal component ",i))
print(pc$loadings[which(abs(pc$loadings[,i])>0.3),i])
}
```
[1] "Principal component 1"
lake_lat_decdeg   avg_temp_degC         avg_ppt   lm_slope_temp 
      0.3941073      -0.4134406      -0.3469947      -0.3169618 
[1] "Principal component 2"
          lake_elevation_m            lake_lon_decdeg         glaciated_category 
                 0.3381356                 -0.3165895                 -0.3620231 
avg_annual_range_temp_degC     avg_nlcd_icesnow12_pct 
                 0.3579748                  0.3790809 
[1] "Principal component 3"
        lake_waterarea_ha          lake_perimeter_m                ws_area_ha lake_lakes1ha_upstream_ha 
               -0.3696102                -0.4118243                -0.3931855                -0.4000449 
 lake_lakes1ha_upstream_n 
               -0.4767508 
[1] "Principal component 4"
 avg_nlcd_forest_pct avg_nlcd_shrub52_pct 
           0.4794306           -0.3426778 
[1] "Principal component 5"
nlcd_summed_agr_diff     avg_nlcd_dev_pct      avg_nlcd_ag_pct 
          -0.4310571           -0.3075739            0.5349932 
[1] "Principal component 6"
nlcd_summed_dev_diff     avg_nlcd_dev_pct avg_nlcd_wetland_pct 
           0.5323473            0.4729988           -0.3476502 
[1] "Principal component 7"
        lake_waterarea_ha          lake_perimeter_m lake_lakes1ha_upstream_ha  lake_lakes1ha_upstream_n 
                0.4565003                 0.4423305                -0.5313871                -0.4397920 
[1] "Principal component 8"
        lm_slope_ppt nlcd_summed_agr_diff avg_nlcd_wetland_pct avg_nlcd_shrub52_pct 
           0.4351944           -0.3185904            0.3135949           -0.5236083 
[1] "Principal component 9"
 nlcd_summed_forest_diff   avg_nlcd_icesnow12_pct avg_nlcd_openwater11_pct 
               0.4571365                0.4219788               -0.4275143 



PCAs overlaid with PCs
```{r}
require(scales)
require(tidyverse)
bg.fui = tibble(
        ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
        ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
        color = paste0(c(
                "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
                "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
                "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04"),"60")
)

bg.fui<-as.data.frame(bg.fui)

       
Y_dwl<- Y %>%
        rownames_to_column() %>%
        rename(lake_nhdid=rowname) %>%
        select(lake_nhdid,avg_dwl_site)

pc.scores <- as.data.frame(pc$scores)
pc.scores<-pc.scores %>% 
        rownames_to_column() %>%
        rename(lake_nhdid=rowname) %>%
        inner_join(.,Y_dwl, by=join_by(lake_nhdid)) %>%
        mutate( col_hex = case_when(avg_dwl_site >=bg.fui[1,1] & avg_dwl_site <bg.fui[1,2] ~ bg.fui[1,3],
                                    avg_dwl_site >=bg.fui[2,1] & avg_dwl_site <bg.fui[2,2] ~bg.fui[2,3], 
                                    avg_dwl_site >=bg.fui[3,1] & avg_dwl_site <bg.fui[3,2] ~bg.fui[3,3]   , 
                                    avg_dwl_site >=bg.fui[4,1] & avg_dwl_site <bg.fui[4,2] ~bg.fui[4,3]   , 
                                    avg_dwl_site >=bg.fui[5,1] & avg_dwl_site <bg.fui[5,2] ~bg.fui[5,3]   , 
                                    avg_dwl_site >=bg.fui[6,1] & avg_dwl_site <bg.fui[6,2] ~bg.fui[6,3]   , 
                                    avg_dwl_site >=bg.fui[7,1] & avg_dwl_site <bg.fui[7,2] ~bg.fui[7,3]   , 
                                    avg_dwl_site >=bg.fui[8,1] & avg_dwl_site <bg.fui[8,2] ~bg.fui[8,3]   , 
                                    avg_dwl_site >=bg.fui[9,1] & avg_dwl_site <bg.fui[9,2] ~bg.fui[9,3]   , 
                                    avg_dwl_site >=bg.fui[10,1] & avg_dwl_site <bg.fui[10,2] ~bg.fui[10,3]   , 
                                    avg_dwl_site >=bg.fui[11,1] & avg_dwl_site <bg.fui[11,2] ~bg.fui[11,3]   , 
                                    avg_dwl_site >=bg.fui[12,1] & avg_dwl_site <bg.fui[12,2] ~bg.fui[12,3]   , 
                                    avg_dwl_site >=bg.fui[13,1] & avg_dwl_site <bg.fui[13,2] ~bg.fui[13,3]   , 
                                    avg_dwl_site >=bg.fui[14,1] & avg_dwl_site <bg.fui[14,2] ~bg.fui[14,3]   , 
                                    avg_dwl_site >=bg.fui[15,1] & avg_dwl_site <bg.fui[15,2] ~bg.fui[15,3]   , 
                                    avg_dwl_site >=bg.fui[16,1] & avg_dwl_site <bg.fui[16,2] ~bg.fui[16,3]   , 
                                    avg_dwl_site >=bg.fui[17,1] & avg_dwl_site <bg.fui[17,2] ~bg.fui[17,3]   , 
                                    avg_dwl_site >=bg.fui[18,1] & avg_dwl_site <bg.fui[18,2] ~bg.fui[18,3]   , 
                                    avg_dwl_site >=bg.fui[19,1] & avg_dwl_site <bg.fui[19,2] ~bg.fui[19,3]   , 
                                    avg_dwl_site >=bg.fui[20,1] & avg_dwl_site <bg.fui[20,2] ~bg.fui[20,3]   , 
                                    avg_dwl_site >=bg.fui[21,1] & avg_dwl_site <=bg.fui[21,2] ~bg.fui[21,3] , T~"black"))


```


Correlation between lake color and PCs
```{r}
lake_color<- pc.scores %>% select(avg_dwl_site, paste0("Comp.",seq(1:27)))
(round(RY<-cor(lake_color)[,1][2:10],digits=2))
(round(RY<-cor(lake_color)[,1][2:28],digits=2)) # only principle component 3 greater than 0.2

# Comp.1 Comp.2 Comp.3 Comp.4 Comp.5 Comp.6 Comp.7 Comp.8 Comp.9 
#  -0.12   0.18  -0.16  -0.61   0.03  -0.05  -0.04  -0.18  -0.02 
# 
#  Comp.1  Comp.2  Comp.3  Comp.4  Comp.5  Comp.6  Comp.7  Comp.8  Comp.9 Comp.10 Comp.11 Comp.12 Comp.13 
#   -0.12    0.18   -0.16   -0.61    0.03   -0.05   -0.04   -0.18   -0.02   -0.08    0.03    0.03    0.02 
# Comp.14 Comp.15 Comp.16 Comp.17 Comp.18 Comp.19 Comp.20 Comp.21 Comp.22 Comp.23 Comp.24 Comp.25 Comp.26 
#   -0.01    0.01    0.02    0.01    0.02   -0.01    0.02   -0.03    0.00    0.03   -0.03    0.07    0.09 
# Comp.27 
#   -0.05
```
NEW!
[1] "Principal component 4"
 avg_nlcd_forest_pct avg_nlcd_shrub52_pct 
           0.4794306           -0.3426778 

Multiple linear regression using a all PCs and combinations of PCs
```{r}
mlm <- lm(avg_dwl_site ~ ., data=lake_color) # all PCs
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site, paste0("Comp.",seq(1:9))) # 1-9 PCs
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(2,4,8)))
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(2,3,4,8)))
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(4)))
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

# [1] 0.02570357
# [1] 0.4753732
# [1] 0.4302942
# [1] 0.4559978
# [1] 0.3674878
```


## FIG 6 - lake color and PC3
```{r}
lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(4)),col_hex)
mlm <- lm(avg_dwl_site ~  Comp.4, data=lake_color)
summary_fit<-summary(mlm)
r_squared <- summary_fit$r.squared
p_value <- summary_fit$coefficients[, 4]

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure6/",Sys.Date(),"_Fig_6_lakecolor_vs_PC4.png"), width = 7, height=5, units = "in", res=400)
par(mar=c(3,3,1,1))
plot(lake_color$Comp.4,lake_color$avg_dwl_site, ylab = "", xlab= "",pch=16, col = lake_color$col_hex, xlim=c(-9,8))
# add equation
text(-9, 495, 
     labels = bquote(y == .(round(summary_fit$coefficients[1,1], 2)) - .(round(abs(summary_fit$coefficients[2,1]), 2)) * x), 
      col = "black", cex = 1.2, adj=0)
# Add R-squared and p-value text
text(-9, 485, 
     labels = bquote(R^2 == .(round(r_squared, 2))), 
        col = "black", cex = 1.2, adj=0)

text(-9, 475, 
     labels = bquote(italic(P) == .(format.pval(p_value, digits = 2))), 
     col = "black", cex = 1.2, adj=0)

abline(lm(avg_dwl_site ~  Comp.4, data=lake_color),lwd=2)
dev.off()
```

## MS stats
```{r}
summary_fit
confint(mlm, level = 0.95)
```
Call:
lm(formula = avg_dwl_site ~ Comp.4, data = lake_color)

Residuals:
    Min      1Q  Median      3Q     Max 
-65.885 -13.932   0.387  12.709 290.616 

Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept) 533.37060 custom_theme    8341   <2e-16 ***
Comp.4       -9.18624    0.04215    -218   <2e-16 ***

Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 18.29 on 81770 degrees of freedom
Multiple R-squared:  0.3675,	Adjusted R-squared:  0.3675 
F-statistic: 4.751e+04 on 1 and 81770 DF,  p-value: < 2.2e-16

                 2.5 %     97.5 %
(Intercept) 533.245271 533.495933
Comp.4       -9.268847  -9.103636



