---
title: "Figures_4-6"
author: "Jordan Von Eggers"
date: "`r Sys.Date()`"
output: html_document
---

```{bash}
ssh jvonegge@beartooth.arcc.uwyo.edu
cd /project/lakecolor/data/

salloc --mem=110GB --nodes=1 --cpus-per-task=8 --account=lakecolor --time=3:00:00

module load arcc/1.0  gcc/12.2.0 
module load r/4.4.0
R
```

```{r}
# Custom theme
  custom_theme <- function() {
  theme_bw() +
    theme(
      text = element_text(color = "black", size = 13),
      axis.text = element_text(color = "black",size = 13),
      axis.title = element_text(color = "black",size = 13),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      legend.text = element_text(size = 13),
      legend.title = element_text(size = 13),
      strip.text = element_text(size = 13, color = "black"))}
```



# Figure 4
## Random forest with training dataset
```{r}
require(randomForest)
require(tidyverse)
require(caret)
library(paletteer)
require(ggplot2)
```


```{r}
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/2024-08-29_lakecolor_climate_landcover_trends.rds") 
nrow(lk) # 81849

# looking at blue vs green lakes
lake_color <- lk %>% 
column_to_rownames(var = "lake_nhdid") %>% 
  mutate(color_category = case_when(
    avg_dwl_site > 530 ~ "Green", 
    avg_dwl_site < 530 ~ "Blue"
  )) %>%
  mutate(color_category = as.factor(color_category)) %>% 
select(        # random forest category
                color_category,
                # lake size and location variables
                epanutr_zoneid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                ws_area_ha,
                # climate variables
                avg_temp_degC,
                avg_ppt,
                lm_slope_temp,
                lm_slope_ppt,
                avg_annual_range_temp_degC,
                # landcover vars
                nlcd_summed_forest_diff,
                nlcd_summed_wetland_diff,
                nlcd_summed_agr_diff,
                nlcd_summed_dev_diff,
                avg_nlcd_dev_pct,
                avg_nlcd_ag_pct,
                avg_nlcd_forest_pct,
                avg_nlcd_wetland_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_barren31_pct,
                avg_nlcd_openwater11_pct
        ) %>% na.omit() # some NAs for watershed area in the TX and GA lakes

nrow(lake_color)
#[1] 81773


# count number of observations for blue and green by eco region
category_counts <- lake_color %>%
  group_by(epanutr_zoneid, color_category) %>% 
  summarise(n = n(), .groups = 'drop')

# subset equally lakes within each ecoregion
set.seed(111)
train <- lake_color %>%
  group_by(epanutr_zoneid) %>%
  sample_frac(.8) %>% #80% of data will be used for training
  ungroup() 
 
validate <- lake_color %>%
   anti_join(train)

# First rF looking at the color category
set.seed(111)
(rF_color <- randomForest(color_category ~ ., data = train, mtry = 5))

# Call:
#  randomForest(formula = color_category ~ ., data = train, mtry = 5) 
#                Type of random forest: classification
#                      Number of trees: 500
# No. of variables tried at each split: 5
# 
#         OOB estimate of  error rate: 14.93%
# Confusion matrix:
#        Blue Green class.error
# Blue  22963  5053   0.1803612 
# Green  4714 32688   0.1260360 

# percent of correct classifications
round((1-0.1803612 )*100,digits=1)
round((1- 0.1260360 )*100,digits=1)

rF_color_pred <- predict(rF_color, validate)
rF_color_pred_confus <-table(observed = validate$color_category, predicted = rF_color_pred)

rm(lake_color);rm(lk)
#save.image(paste0(Sys.Date(),"_random_forest_color.Rdata"))

load("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Code/Data_Figures_4-6/2024-08-30_random_forest_color.Rdata")


cm <- confusionMatrix(rF_color_pred,validate$color_category, dnn=c('Predicted', 'Observed'))

accuracy <- cm$overall

cm <- as.data.frame(cm$table) |>
  group_by(Observed) |>
  mutate(prop = Freq/sum(Freq),
         prop = round(prop,2))


ggplot(cm, aes(x = Observed, y = Predicted, fill=Freq)) +
  geom_tile(color="black") +
  scale_x_discrete(expand = c(0, 0))+ #remove white space
  scale_y_discrete(expand = c(0, 0))+ #remove white space
  scale_fill_gradient(low="white", high="orangered",
                      name="Frequency") +
  geom_text(aes(label = paste0("n=",Freq)), vjust = .5,  alpha = 1, size=5) +
  geom_text(aes(label = paste0(prop*100,"%")), vjust = 2.0,  alpha = 1, size=4) +
  custom_theme() +
  theme(legend.position = 'none') +
  labs(title = paste0("Accuracy: ",round(accuracy,2))) 

ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/CM_rF_color.png",width=3.7, height=2.5, units="in",dpi = 300)




# create prettier lollipop plot for the color 
rF_imp<-as.data.frame(importance(rF_color)) %>%
       rownames_to_column(.,var="variables")

# update names, fist print as the string that is easy to edit
(paste0("c(", paste0('"', rF_imp$variables, '"', collapse = ", "), ")"))

rF_imp$newvarnames<-c("Ecoregion", "Lake elevation (m)", "Latitude (DD)", "Longitude (DD)", "Lake area (ha)", "Lake perimeter (m)", "Watershed area (ha)", "Average temperature (deg C)", "Average precipitation (ppt)", "Slope of temperature over time", "Slope of precipitation over time", "Average annual temperature range (deg C)", "Difference in forest 2001-2016 (%)", "Difference in wetland 2001-2016 (%)", "Difference in agriculture 2001-2016 (%)", "Difference in development 2001-2016 (%)", "Average development (%)", "Average agriculture (%)", "Average forest (%)", "Average wetland (%)", "Average ice and snow (%)", "Average shrub (%)", "Average barren (%)", "Average open water (%)")

rF_imp <- rF_imp %>% arrange(MeanDecreaseGini) %>%  mutate(newvarnames = factor(newvarnames, levels = newvarnames)) 
write.csv(rF_imp, "/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Code/Data_Figures_4-6/rF_importance_color.csv")

rF_imp <- rF_imp %>% slice_tail(n=10) # just plot the top 10


ggplot(rF_imp, aes(y = MeanDecreaseGini, x = newvarnames)) +
  geom_segment(aes(x = newvarnames, xend = newvarnames, y = 0, yend = MeanDecreaseGini), color = "black") +
  geom_point(aes(color = MeanDecreaseGini), size = 4) +
  scale_color_paletteer_c("grDevices::Zissou 1", n = 30, direction = 1) +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
            panel.grid.minor.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(), 
    axis.text = element_text(color = "black", size = 15),  
    axis.title = element_text(color = "black", size = 15), 
    legend.position = "none",
    plot.margin = unit(c(0.2, 0.6, 0.2, 0.2), "cm")) + 
  xlab("") + 
  ylab("Mean decrease in Gini")


ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/Fig4_Importance_rF_color.png",width=6.7, height=4, units="in",dpi = 300)
```
total number of lakes:
[1] 81773

```{r}
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/2024-08-29_lakecolor_climate_landcover_trends.rds") 
nrow(lk) 


# looking at basic trend categories (trending blue, trending green, not trending)
lake_color <- lk %>% 
column_to_rownames(var = "lake_nhdid") %>% 
mutate(basic_trend_cat=case_when(lm_slope_color > 0 & lm_pval_color <= 0.05~ "Trending green", 
                              lm_slope_color < 0 & lm_pval_color <= 0.05~ "Trending blue", 
                              lm_pval_color > 0.05~ "No trend")) %>%
        mutate(basic_trend_cat=as.factor(basic_trend_cat)) %>% 
select(        # random forest category
               basic_trend_cat,
                # lake size and location variables
                epanutr_zoneid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                ws_area_ha,
                # climate variables
                avg_temp_degC,
                avg_ppt,
                lm_slope_temp,
                lm_slope_ppt,
                avg_annual_range_temp_degC,
                # landcover vars
                nlcd_summed_forest_diff,
                nlcd_summed_wetland_diff,
                nlcd_summed_agr_diff,
                nlcd_summed_dev_diff,
                avg_nlcd_dev_pct,
                avg_nlcd_ag_pct,
                avg_nlcd_forest_pct,
                avg_nlcd_wetland_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_barren31_pct,
                avg_nlcd_openwater11_pct
        ) %>% na.omit() # some NAs for watershed area in the TX and GA lakes

nrow(lake_color)
#[1] 81773

# count number of observations for blue and green and no trend by eco region
category_counts <- lake_color %>%
  group_by(epanutr_zoneid, basic_trend_cat) %>% 
  summarise(n = n(), .groups = 'drop')

# what is the minimum number of observations of one of the categories to subset by
min(category_counts$n)
#341

# sample 340 lakes from each ecoregion to make the number of trend categories even
data_subset <- lake_color %>%
  group_by(epanutr_zoneid, basic_trend_cat) %>%
  slice_sample(n = 340, replace = FALSE)  


set.seed(111)
train <- data_subset %>%
  group_by(epanutr_zoneid) %>%
  sample_frac(.8) %>% #80% of data will be used for training
  ungroup() 
 
validate <- data_subset %>%
   anti_join(train)

set.seed(111)
(rF_trend<-randomForest(basic_trend_cat~.,data=train, mtry=5))

# Call:
#  randomForest(formula = basic_trend_cat ~ ., data = train, mtry = 5) 
#                Type of random forest: classification
#                      Number of trees: 500
# No. of variables tried at each split: 5
# 
#         OOB estimate of  error rate: 53.62%
# Confusion matrix:
#                No trend Trending blue Trending green class.error
# No trend            888           829            731   0.6372549 
# Trending blue       650          1335            469   0.4559902 
# Trending green      659           600           1183   0.5155610 

# percent of correct classifications
round((1-0.6372549)*100,digits=1)
round((1-0.4559902)*100,digits=1)
round((1-0.5155610)*100,digits=1)


rm(lake_color);rm(lk)
#save.image(paste0(Sys.Date(),"_random_forest_trend.Rdata"))

load("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Code/Data_Figures_4-6/2024-08-30_random_forest_trend.Rdata")

rF_trend_pred <- predict(rF_trend, validate)
rF_trend_pred_confus <-table(observed =  validate$basic_trend_cat, predicted = rF_trend_pred)

cm <- confusionMatrix(rF_trend_pred,validate$basic_trend_cat, dnn=c('Predicted', 'Observed'))

accuracy <- cm$overall

cm <- as.data.frame(cm$table) |>
  group_by(Observed) |>
  mutate(prop = Freq/sum(Freq),
         prop = round(prop,2))



ggplot(cm, aes(x = Observed, y = Predicted, fill=Freq)) +
  geom_tile(color="black") +
  scale_x_discrete(expand = c(0, 0))+ #remove white space
  scale_y_discrete(expand = c(0, 0))+ #remove white space
  scale_fill_gradient(low="white", high="orangered",
                      name="Frequency") +
  geom_text(aes(label = paste0("n=",Freq)), vjust = .5,  alpha = 1, size=5) +
  geom_text(aes(label = paste0(prop*100,"%")), vjust = 2.0,  alpha = 1, size=4) +
  custom_theme() +
  theme(legend.position = 'none') +
  labs(title = paste0("Accuracy: ",round(accuracy,2))) 

ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/CM_rF_trend.png",width=6, height=2.5, units="in",dpi = 300)



# create prettier lollipop plot 
rF_imp<-as.data.frame(importance(rF_trend)) %>%
       rownames_to_column(.,var="variables")

# update names, fist print as the string that is easy to edit
(paste0("c(", paste0('"', rF_imp$variables, '"', collapse = ", "), ")"))

rF_imp$newvarnames<-c("Ecoregion", "Lake elevation (m)", "Latitude (DD)", "Longitude (DD)", "Lake area (ha)", "Lake perimeter (m)", "Watershed area (ha)", "Average temperature (deg C)", "Average precipitation (ppt)", "Slope of temperature over time", "Slope of precipitation over time", "Average annual temperature range (deg C)", "Difference in forest 2001-2016 (%)", "Difference in wetland 2001-2016 (%)", "Difference in agriculture 2001-2016 (%)", "Difference in development 2001-2016 (%)", "Average development (%)", "Average agriculture (%)", "Average forest (%)", "Average wetland (%)", "Average ice and snow (%)", "Average shrub (%)", "Average barren (%)", "Average open water (%)")

rF_imp <- rF_imp %>% arrange(MeanDecreaseGini) %>%  mutate(newvarnames = factor(newvarnames, levels = newvarnames)) 
write.csv(rF_imp, "/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Code/Data_Figures_4-6/rF_importance_trend.csv")

rF_imp <- rF_imp %>% slice_tail(n=10) # just plot the top 10


ggplot(rF_imp, aes(y = MeanDecreaseGini, x = newvarnames)) +
  geom_segment(aes(x = newvarnames, xend = newvarnames, y = 0, yend = MeanDecreaseGini), color = "black") +
  geom_point(aes(color = MeanDecreaseGini), size = 4) +
scale_color_paletteer_c("grDevices::Zissou 1", n = 30, direction=1) +
        theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(), 
axis.text = element_text(color = "black", size=15),  
    axis.title = element_text(color = "black",size=15),  
     legend.position = "none",
 plot.margin = unit(c(0.2, 0.6, 0.2, 0.2), "cm")) + 
  xlab("") + 
  ylab("Mean decrease in Gini") 


ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/Fig4_Importance_rF_trend.png",width=6.7, height=4, units="in",dpi = 300)


```


### Plot Random Forest
```{r}
load("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/RandomForest/2024-06-27_random_forest_output_crossval_train.Rdata")

require(randomForest)
library(paletteer)
require(ggplot2)
```

### Lollipop RF color and trends
```{r}
# create prettier lollipop plot 
rF_imp<-as.data.frame(importance(rF_trend)) %>%
       rownames_to_column(.,var="variables")

# update names, fist print as the string that is easy to edit
(paste0("c(", paste0('"', rF_imp$variables, '"', collapse = ", "), ")"))

rF_imp$newvarnames<-c("Ecoregion", "Lake elevation (m)", "Latitude (DD)", "Longitude (DD)", "Lake area (ha)", "Lake perimeter (m)", "Watershed area (ha)", "Average temperature (deg C)", "Average precipitation (ppt)", "Slope of temperature over time", "Slope of precipitation over time", "Average annual temperature range (deg C)", "Difference in forest 2001-2016 (%)", "Difference in wetland 2001-2016 (%)", "Difference in agriculture 2001-2016 (%)", "Difference in development 2001-2016 (%)", "Average development (%)", "Average agriculture (%)", "Average forest (%)", "Average wetland (%)", "Average ice and snow (%)", "Average shrub (%)", "Average barren (%)", "Average open water (%)")

rF_imp <- rF_imp %>% arrange(MeanDecreaseGini) %>%  mutate(newvarnames = factor(newvarnames, levels = newvarnames)) %>% slice_tail(n=10)


ggplot(rF_imp, aes(y = MeanDecreaseGini, x = newvarnames)) +
  geom_segment(aes(x = newvarnames, xend = newvarnames, y = 0, yend = MeanDecreaseGini), color = "black") +
  geom_point(aes(color = MeanDecreaseGini), size = 4) +
scale_color_paletteer_c("grDevices::Zissou 1", n = 30, direction=1) +
        theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(), 
axis.text = element_text(color = "black", size=15),  
    axis.title = element_text(color = "black",size=15),  
     legend.position = "none",
 plot.margin = unit(c(0.2, 0.6, 0.2, 0.2), "cm")) + 
  xlab("") + 
  ylab("Mean decrease in Gini") 


ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/Fig4_Importance_rF_trend.png",width=6.7, height=4, units="in",dpi = 300)

# create prettier lollipop plot for the color 
rF_imp<-as.data.frame(importance(rF_color)) %>%
       rownames_to_column(.,var="variables")

# update names, fist print as the string that is easy to edit
(paste0("c(", paste0('"', rF_imp$variables, '"', collapse = ", "), ")"))

rF_imp$newvarnames<-c("Ecoregion", "Lake elevation (m)", "Latitude (DD)", "Longitude (DD)", "Lake area (ha)", "Lake perimeter (m)", "Watershed area (ha)", "Average temperature (deg C)", "Average precipitation (ppt)", "Slope of temperature over time", "Slope of precipitation over time", "Average annual temperature range (deg C)", "Difference in forest 2001-2016 (%)", "Difference in wetland 2001-2016 (%)", "Difference in agriculture 2001-2016 (%)", "Difference in development 2001-2016 (%)", "Average development (%)", "Average agriculture (%)", "Average forest (%)", "Average wetland (%)", "Average ice and snow (%)", "Average shrub (%)", "Average barren (%)", "Average open water (%)")

rF_imp <- rF_imp %>% arrange(MeanDecreaseGini) %>%  mutate(newvarnames = factor(newvarnames, levels = newvarnames)) %>% slice_tail(n=10)


ggplot(rF_imp, aes(y = MeanDecreaseGini, x = newvarnames)) +
  geom_segment(aes(x = newvarnames, xend = newvarnames, y = 0, yend = MeanDecreaseGini), color = "black") +
  geom_point(aes(color = MeanDecreaseGini), size = 4) +
  scale_color_paletteer_c("grDevices::Zissou 1", n = 30, direction = 1) +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
            panel.grid.minor.x = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(), 
    axis.text = element_text(color = "black", size = 15),  
    axis.title = element_text(color = "black", size = 15), 
    legend.position = "none",
    plot.margin = unit(c(0.2, 0.6, 0.2, 0.2), "cm")) + 
  xlab("") + 
  ylab("Mean decrease in Gini")


ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure4/Fig4_Importance_rF_color.png",width=6.7, height=4, units="in",dpi = 300)
```

### Color confusion matrix
```{r}
require(caret)
load("2024-05-16_regression_data.Rdata")


# looking at blue vs green lakes
X_N <- rownames_to_column(X, var = "rownames")
Y_N <- rownames_to_column(Y, var = "rownames")
lake_color <- inner_join(X_N,Y_N,by="rownames") %>% 
        select(-c(trend_cat,trend_cat_switch,lm_slope_color,lm_pval_color)) %>%
               column_to_rownames(var = "rownames") %>% 
        mutate(color_category=case_when(avg_dwl_site > 530 ~ "Green", 
                              avg_dwl_site < 530 ~ "Blue")) %>%
        mutate(color_category=as.factor(color_category)) %>% 
        select(-avg_dwl_site, -mean_range_dwl_site)


set.seed(111)
ind <- sample(2, nrow(lake_color), replace = TRUE, prob=c(0.8, 0.2))

# First rF lookinag at the color category
set.seed(111)
(rF_color<-randomForest(color_category~.,data=lake_color[ind == 1,], mtry=5))
rF_color_pred <- predict(rF_color, lake_color[ind == 2,])

rF_color_pred_confus <-table(observed = lake_color[ind==2, "color_category"], predicted = rF_color_pred)


require(caret)
cm <- confusionMatrix(rF_color_pred,lake_color[ind==2, "color_category"], dnn=c('Predicted', 'Observed'))

accuracy <- cm$overall

cm <- as.data.frame(cm$table) |>
  group_by(Observed) |>
  mutate(prop = Freq/sum(Freq),
         prop = round(prop,2))



ggplot(cm, aes(x = Observed, y = Predicted, fill=Freq)) +
  geom_tile(color="black") +
  scale_x_discrete(expand = c(0, 0))+ #remove white space
  scale_y_discrete(expand = c(0, 0))+ #remove white space
  scale_fill_gradient(low="white", high="orangered",
                      name="Frequency") +
  geom_text(aes(label = paste0("n=",Freq)), vjust = .5,  alpha = 1, size=5) +
  geom_text(aes(label = paste0(prop*100,"%")), vjust = 2.0,  alpha = 1, size=4) +
  custom_theme() +
  theme(legend.position = 'none') +
  labs(title = paste0("Accuracy: ",round(accuracy,2))) 

ggsave("Figure4/CM_rF_color.png",width=3.7, height=2.5, units="in",dpi = 300)


```


### Trend confusion matrix
```{r}

# looking at basic trend categories (trending blue, trending green, not trending)
lake_color <- inner_join(X_N,Y_N,by="rownames") %>% 
        select(-c(trend_cat,trend_cat_switch,avg_dwl_site )) %>%
               column_to_rownames(var = "rownames") %>% 
        select(lm_slope_color,lm_pval_color, everything()) %>%
        mutate(basic_trend_cat=case_when(lm_slope_color > 0 & lm_pval_color <= 0.05~ "Trending green", 
                              lm_slope_color < 0 & lm_pval_color <= 0.05~ "Trending blue", 
                              lm_pval_color > 0.05~ "No trend")) %>%
        mutate(basic_trend_cat=as.factor(basic_trend_cat)) %>%
        select(-c(lm_pval_color, lm_slope_color,mean_range_dwl_site))

# without including the variability in DWL

set.seed(111)
(rF_basic_norange<-randomForest(basic_trend_cat~.,data=lake_color[ind == 1,], mtry=5))
rF_basic_norange_pred <- predict(rF_basic_norange, lake_color[ind == 2,])
rF_basic_norange_pred_confus <-table(observed = lake_color[ind==2, "basic_trend_cat"], predicted = rF_basic_norange_pred)


require(caret)
cm <- confusionMatrix(rF_basic_norange_pred,lake_color[ind==2, "basic_trend_cat"], dnn=c('Predicted', 'Observed'))

accuracy <- cm$overall

cm <- as.data.frame(cm$table) |>
  group_by(Observed) |>
  mutate(prop = Freq/sum(Freq),
         prop = round(prop,2))



ggplot(cm, aes(x = Observed, y = Predicted, fill=Freq)) +
  geom_tile(color="black") +
  scale_x_discrete(expand = c(0, 0))+ #remove white space
  scale_y_discrete(expand = c(0, 0))+ #remove white space
  scale_fill_gradient(low="white", high="orangered",
                      name="Frequency") +
  geom_text(aes(label = paste0("n=",Freq)), vjust = .5,  alpha = 1, size=5) +
  geom_text(aes(label = paste0(prop*100,"%")), vjust = 2.0,  alpha = 1, size=4) +
  custom_theme() +
  theme(legend.position = 'none') +
  labs(title = paste0("Accuracy: ",round(accuracy,2))) 

ggsave("Figure4/CM_rF_trend.png",width=6, height=2.5, units="in",dpi = 300)
```


```{r}
# Custom theme
  custom_theme <- function() {
  theme_bw() +
    theme(
      text = element_text(color = "black", size = 13),
      axis.text = element_text(color = "black",size = 13),
      axis.title = element_text(color = "black",size = 13),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12),
      legend.text = element_text(size = 13),
      legend.title = element_text(size = 13),
      strip.text = element_text(size = 13, color = "black"))}
```




# Figure 5


```{r}
require(tidyverse)

lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/2024-08-29_lakecolor_climate_landcover_trends.rds")
nrow(lk)
#[1] 81874

# #filter lakes greater than 100,000 hectares
# lk <- lk %>% filter(lake_waterarea_ha <= 100000) 
# nrow(lk)
# #[1] 81868 # removed 6 lakes greater than 100,00 hectares

# remove lakes with wavelengths greater than 590
lk <- lk %>% filter(avg_dwl_site < 590) 

# looking at blue vs green lakes
lake_color <- lk %>% select(-c(trend_cat, trend_cat_switch, lm_slope_color, lm_pval_color)) %>% 
  mutate(color_category = case_when(
    avg_dwl_site > 530 ~ "Green", 
    avg_dwl_site < 530 ~ "Blue"
  )) %>%
  mutate(color_category = as.factor(color_category)) %>% 
  select(-avg_dwl_site, -mean_range_dwl_site) #%>%
 # filter(lake_waterarea_ha <= 100000)

# Log transform lake area, lake perimeter and watershed area
lake_color<- lake_color %>%
        mutate(lake_waterarea_ha=log(lake_waterarea_ha),
               ws_area_ha=log(ws_area_ha),
               lake_perimeter_m=log(lake_perimeter_m))

# Reshape data to long format
lake_color_long <- lake_color %>%
  pivot_longer(cols = c("avg_nlcd_forest_pct", "lake_lon_decdeg", "lake_lat_decdeg", "avg_temp_degC", "ws_area_ha", "lake_elevation_m", "lake_waterarea_ha", "lake_perimeter_m", "avg_annual_range_temp_degC", "avg_nlcd_openwater11_pct"), 
               names_to = "variable", 
               values_to = "value")

# Create a named vector for renaming the variables
new_names <- c(
  "avg_nlcd_forest_pct" = "Average forest (%)",
  "lake_lon_decdeg" = "Longitude (DD)",
  "lake_lat_decdeg" = "Latitude (DD)",
  "avg_temp_degC" = "Average temperature (deg C)",
  "ws_area_ha" = "log(Watershed area (ha))",
  "lake_elevation_m" = "Lake elevation (m)",
  "lake_waterarea_ha" = "log(Lake area (ha))",
  "lake_perimeter_m" = "log(Lake perimeter (m))",
  "avg_annual_range_temp_degC" = "Average annual\ntemperature range (deg C)",
  "avg_nlcd_openwater11_pct" = "Average open water (%)"
)

# Rename the variables using recode
lake_color_long <- lake_color_long %>%
  mutate(variable = recode(variable, !!!new_names))



# Create density plots
plt<-ggplot(lake_color_long, aes(x = value, fill = color_category)) +
  geom_density(alpha = 0.7) +
  facet_wrap(~ variable, nrow=2,scales = "free") +
  labs(x = "",
       y = "Density",
       fill="Lake color") +
          scale_fill_manual(values = c("Green" = "#759e72", "Blue"="#327cbb")) +
  theme_classic() +
  theme(legend.position = "top") 
        

ggsave("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure5/Figure_5_Density_plots_important_vars.png",width=12, height=6, units="in",dpi = 200)
```




## MS stats
```{r}

require(tidyverse)

lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/2024-08-29_lakecolor_climate_landcover_trends.rds")
nrow(lk)
#[1] 81874

# #filter lakes greater than 100,000 hectares
# lk <- lk %>% filter(lake_waterarea_ha <= 100000) 
# nrow(lk)
# #[1] 81868 # removed 6 lakes greater than 100,00 hectares

# remove lakes with wavelengths greater than 590
lk <- lk %>% filter(avg_dwl_site < 590) 

# looking at blue vs green lakes
lake_color <- lk %>% select(-c(trend_cat, trend_cat_switch, lm_slope_color, lm_pval_color)) %>% 
  mutate(color_category = case_when(
    avg_dwl_site > 530 ~ "Green", 
    avg_dwl_site < 530 ~ "Blue"
  )) %>%
  mutate(color_category = as.factor(color_category)) %>% 
  select(-avg_dwl_site, -mean_range_dwl_site) #%>%
 # filter(lake_waterarea_ha <= 100000)

# # Log transform lake area, lake perimeter and watershed area
# lake_color<- lake_color %>%
#         mutate(lake_waterarea_ha=log(lake_waterarea_ha),
#                ws_area_ha=log(ws_area_ha),
#                lake_perimeter_m=log(lake_perimeter_m))

# Reshape data to long format
lake_color_long <- lake_color %>%
  pivot_longer(cols = c("avg_nlcd_forest_pct", "lake_lon_decdeg", "lake_lat_decdeg", "avg_temp_degC", "ws_area_ha", "lake_elevation_m", "lake_waterarea_ha", "lake_perimeter_m", "avg_annual_range_temp_degC", "avg_nlcd_openwater11_pct"), 
               names_to = "variable", 
               values_to = "value")

# Create a named vector for renaming the variables
new_names <- c(
  "avg_nlcd_forest_pct" = "Average forest (%)",
  "lake_lon_decdeg" = "Longitude (DD)",
  "lake_lat_decdeg" = "Latitude (DD)",
  "avg_temp_degC" = "Average temperature (deg C)",
  "ws_area_ha" = "Watershed area (ha)",
  "lake_elevation_m" = "Lake elevation (m)",
  "lake_waterarea_ha" = "Lake area (ha)",
  "lake_perimeter_m" = "Lake perimeter (m)",
  "avg_annual_range_temp_degC" = "Average annual\ntemperature range (deg C)",
  "avg_nlcd_openwater11_pct" = "Average open water (%)"
)

# Rename the variables using recode
lake_color_long <- lake_color_long %>%
  mutate(variable = recode(variable, !!!new_names))


summary_stats <- lake_color_long %>%
  group_by(variable, color_category) %>% 
  summarise(
    mean_value = round(mean(value, na.rm = TRUE),digits=2),      
    sd_value = round(sd(value, na.rm = TRUE), digits=2)          
  )



# Assuming your dataframe is named lake_color_long, with columns: value, variable, and color_category
t_test_results <- lake_color_long %>%
  group_by(variable) %>% # Group by variable
  summarise(
    t_test = list(t.test(value ~ color_category, data = pick(value, color_category))) # Specify columns in pick()
  ) %>%
  mutate(
    t_statistic = sapply(t_test, function(x) x$statistic),   # Extract t-statistic
    p_value = sapply(t_test, function(x) x$p.value),         # Extract p-value
    conf_low = sapply(t_test, function(x) x$conf.int[1]),    # Extract confidence interval lower bound
    conf_high = sapply(t_test, function(x) x$conf.int[2])    # Extract confidence interval upper bound
  ) %>%
  select(-t_test) %>%  # Remove the list column if you just want the summary statistics 
        mutate(pval_lessthan0.5=ifelse(p_value<0.05, "Yes","No"))
```



# Figure 6
## PCA
```{r}
require(tidyverse)
lk<-readRDS("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Data/2024-08-29_lakecolor_climate_landcover_trends.rds")
nrow(lk)
#[1] 81849

# select the variables you want to work on, but it will automatically grab date and lake_ndhid
X <- lk %>%
        select(
                # lake size and location variables
                lake_nhdid,
                lake_elevation_m,
                lake_lat_decdeg,
                lake_lon_decdeg,
                lake_waterarea_ha,
                lake_perimeter_m,
                ws_area_ha,
                # climate variables
                avg_temp_degC,
                avg_ppt,
                lm_slope_temp,
                lm_slope_ppt,
                avg_annual_range_temp_degC,
                # landcover vars
                nlcd_summed_forest_diff,
                nlcd_summed_wetland_diff,
                nlcd_summed_agr_diff,
                nlcd_summed_dev_diff,
                avg_nlcd_dev_pct,
                avg_nlcd_ag_pct,
                avg_nlcd_forest_pct,
                avg_nlcd_wetland_pct,
                avg_nlcd_icesnow12_pct,
                avg_nlcd_shrub52_pct,
                avg_nlcd_barren31_pct,
                avg_nlcd_openwater11_pct
        )
dim(X)
#[1]  81849    24

X <- X %>%
       column_to_rownames(var = "lake_nhdid") %>% distinct() %>% drop_na()
dim(X)
#[1] 81773    23

Y <- lk %>%
               filter(lake_nhdid%in%rownames(X))%>%
        select(lake_nhdid,avg_dwl_site, trend_cat, trend_cat_switch, lm_slope_color, lm_pval_color)
 
Y <- Y %>%
       column_to_rownames(var = "lake_nhdid") %>% distinct() %>% drop_na()
dim(Y)
#[1] 81772     5

X <- X %>% filter(rownames(.)%in%rownames(Y))
dim(X)
#[1]  81772    23

rm(list=ls()[! ls() %in% c("Y", "X")])

# run PCA
R <- cor(X)
round(R,4)

pc <- princomp(X, cor=T)
```


```{r}
(spc <- summary(pc, loadings=T, cutoff=0.3))
#lambda <- pc$sd^2
plot(pc, type="lines", main = "")
```
Keep 8 PCs

```{r}
for (i in 1:8){
print(paste0("Principal component ",i))
print(pc$loadings[which(abs(pc$loadings[,i])>0.3),i])
}
```
[1] "Principal component 1"
lake_elevation_m  lake_lat_decdeg  lake_lon_decdeg 
       0.3091502        0.3555449       -0.3020105 
   avg_temp_degC          avg_ppt 
      -0.3834401       -0.3549859 
[1] "Principal component 2"
      lake_elevation_m          lm_slope_temp 
             0.3479156              0.3955738 
   avg_nlcd_forest_pct avg_nlcd_icesnow12_pct 
             0.3096396              0.3701239 
[1] "Principal component 3"
avg_annual_range_temp_degC        avg_nlcd_forest_pct 
                -0.3600478                  0.4346655 
  avg_nlcd_openwater11_pct 
                -0.3145581 
[1] "Principal component 4"
lake_waterarea_ha  lake_perimeter_m        ws_area_ha 
       -0.5487974        -0.5852858        -0.4413294 
[1] "Principal component 5"
nlcd_summed_agr_diff     avg_nlcd_dev_pct      avg_nlcd_ag_pct 
           0.3999372            0.3594142           -0.5176130 
[1] "Principal component 6"
nlcd_summed_dev_diff     avg_nlcd_dev_pct avg_nlcd_wetland_pct 
           0.4897053            0.4236186           -0.3664174 
[1] "Principal component 7"
        lm_slope_ppt nlcd_summed_agr_diff avg_nlcd_wetland_pct 
           0.4342863           -0.3126998            0.3114225 
avg_nlcd_shrub52_pct 
          -0.5396761 
[1] "Principal component 8"
 nlcd_summed_forest_diff   avg_nlcd_icesnow12_pct 
               0.4281223                0.4102663 
   avg_nlcd_barren31_pct avg_nlcd_openwater11_pct 
               0.3070859               -0.4424409 



PCAs overlaid with PCs
```{r}
require(scales)
require(tidyverse)
bg.fui = tibble(
        ymin = c(470,475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583),
        ymax = c(475,480,485,489,495,509,530,549,559,564,567,568,569,570,573,575,577,579,581,583,590),
        color = paste0(c(
                "#2158bc", "#316dc5", "#327cbb", "#4b80a0", "#568f96", "#6d9298", "#698c86", 
                "#759e72", "#7ba654", "#7dae38", "#94b660","#94b660", "#a5bc76", "#aab86d", 
                "#adb55f", "#a8a965", "#ae9f5c", "#b3a053", "#af8a44", "#a46905", "#9f4d04"),"60")
)

bg.fui<-as.data.frame(bg.fui)

       
Y_dwl<- Y %>%
        rownames_to_column() %>%
        rename(lake_nhdid=rowname) %>%
        select(lake_nhdid,avg_dwl_site)

pc.scores <- as.data.frame(pc$scores)
pc.scores<-pc.scores %>% 
        rownames_to_column() %>%
        rename(lake_nhdid=rowname) %>%
        inner_join(.,Y_dwl, by=join_by(lake_nhdid)) %>%
        mutate( col_hex = case_when(avg_dwl_site >=bg.fui[1,1] & avg_dwl_site <bg.fui[1,2] ~ bg.fui[1,3],
                                    avg_dwl_site >=bg.fui[2,1] & avg_dwl_site <bg.fui[2,2] ~bg.fui[2,3], 
                                    avg_dwl_site >=bg.fui[3,1] & avg_dwl_site <bg.fui[3,2] ~bg.fui[3,3]   , 
                                    avg_dwl_site >=bg.fui[4,1] & avg_dwl_site <bg.fui[4,2] ~bg.fui[4,3]   , 
                                    avg_dwl_site >=bg.fui[5,1] & avg_dwl_site <bg.fui[5,2] ~bg.fui[5,3]   , 
                                    avg_dwl_site >=bg.fui[6,1] & avg_dwl_site <bg.fui[6,2] ~bg.fui[6,3]   , 
                                    avg_dwl_site >=bg.fui[7,1] & avg_dwl_site <bg.fui[7,2] ~bg.fui[7,3]   , 
                                    avg_dwl_site >=bg.fui[8,1] & avg_dwl_site <bg.fui[8,2] ~bg.fui[8,3]   , 
                                    avg_dwl_site >=bg.fui[9,1] & avg_dwl_site <bg.fui[9,2] ~bg.fui[9,3]   , 
                                    avg_dwl_site >=bg.fui[10,1] & avg_dwl_site <bg.fui[10,2] ~bg.fui[10,3]   , 
                                    avg_dwl_site >=bg.fui[11,1] & avg_dwl_site <bg.fui[11,2] ~bg.fui[11,3]   , 
                                    avg_dwl_site >=bg.fui[12,1] & avg_dwl_site <bg.fui[12,2] ~bg.fui[12,3]   , 
                                    avg_dwl_site >=bg.fui[13,1] & avg_dwl_site <bg.fui[13,2] ~bg.fui[13,3]   , 
                                    avg_dwl_site >=bg.fui[14,1] & avg_dwl_site <bg.fui[14,2] ~bg.fui[14,3]   , 
                                    avg_dwl_site >=bg.fui[15,1] & avg_dwl_site <bg.fui[15,2] ~bg.fui[15,3]   , 
                                    avg_dwl_site >=bg.fui[16,1] & avg_dwl_site <bg.fui[16,2] ~bg.fui[16,3]   , 
                                    avg_dwl_site >=bg.fui[17,1] & avg_dwl_site <bg.fui[17,2] ~bg.fui[17,3]   , 
                                    avg_dwl_site >=bg.fui[18,1] & avg_dwl_site <bg.fui[18,2] ~bg.fui[18,3]   , 
                                    avg_dwl_site >=bg.fui[19,1] & avg_dwl_site <bg.fui[19,2] ~bg.fui[19,3]   , 
                                    avg_dwl_site >=bg.fui[20,1] & avg_dwl_site <bg.fui[20,2] ~bg.fui[20,3]   , 
                                    avg_dwl_site >=bg.fui[21,1] & avg_dwl_site <=bg.fui[21,2] ~bg.fui[21,3] , T~"black"))


# rename variable labels


rownames(pc$loadings) <- c(
        "Elevation (m)",
        "Latitude",
        "Longitutde",
        "Lake area (ha)",
        "Lake perimeter (m)",
        "W.S. area (ha)",
        "Temp (C)",
        "Precip (ppt)",
        "Temp slope",
        "Precip slope",
        "Range temp (C)",
        "Forest diff",
        "Wetland diff",
        "Ag diff",
        "Dev diff",
        "Dev (%)",
        "Ag (%)",
        "Forest (%)",
        "Wetland (%)",
        "Icesnow (%)",
        "Shrub (%)",
        "Barren (%)",
        "Openwater (%)"
)




i=1
for(i in 1:8){
load <- pc$loadings[, c(i,i+1)]
load <- load[abs(load[,1])>0.3 | abs(load[,2])>0.3, ]

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure6/PCA/",Sys.Date(),"_PC",i,"_PC",i+1,"_SepScale.png"), width = 8, height=6, units = "in", res=400)
par(mar=c(5,5,1,1), xpd=T)
plot(pc.scores[,i+1], pc.scores[,i+2], pch = 19, col = pc.scores$col_hex, xlab=paste0("PC",i),ylab=paste0("PC",i+1))

# scaling factor for arrow length
scaling_factor <- c(11,10,10,10,10,10,10,10)[i]


# Add arrows for loadings of PC1 and PC2
arrows(0, 0, load[, 1] * scaling_factor, load[, 2] * scaling_factor, 
       angle = 20, length = 0.1, col = "black")


text(load[, 1] * scaling_factor, load[, 2] * scaling_factor,
     labels = rownames(load), pos = 3, col = "black", cex = 1.1)
dev.off()
}

rm(load)
rm(scaling_factor)
rm(bg.fui)
rm(i)
```


Correlation between lake color and PCs
```{r}
lake_color<- pc.scores %>% select(avg_dwl_site, paste0("Comp.",seq(1:23)))
(round(RY<-cor(lake_color)[,1][2:9],digits=2))
(round(RY<-cor(lake_color)[,1][2:24],digits=2)) # only principle component 3 greater than 0.2

# Comp.1 Comp.2 Comp.3 Comp.4 Comp.5 Comp.6 Comp.7 Comp.8 
#  -0.04  -0.11  -0.62   0.05  -0.03  -0.10  -0.19  -0.02
```
Component 3 (avg_annual_range_temp_degC, avg_nlcd_forest_pct, avg_nlcd_openwater11_pct) has the highest correlation with lake color, and no other principal components (components 1-23) have a correlation higher than 0.2 

 

Multiple linear regression using a all PCs and combinations of PCs
```{r}
mlm <- lm(avg_dwl_site ~ ., data=lake_color) # all PCs
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site, paste0("Comp.",seq(1:9))) # 1-9 PCs
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(3,6,7)))
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(3,2,6,7)))
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(3)))
mlm <- lm(avg_dwl_site ~ ., data=lake_color)
summary(mlm)$r.squared

# [1] 0.487073
# [1] 0.4625903
# [1] 0.421627
# [1] 0.4326043
# [1] 0.3743985
```


## FIG 6 - lake color and PC3
```{r}
lake_color <- pc.scores %>% select(avg_dwl_site,  paste0("Comp.",c(3)),col_hex)
mlm <- lm(avg_dwl_site ~  Comp.3, data=lake_color)
summary_fit<-summary(mlm)
#Adjusted R-squared:  0.3745
r_squared <- summary_fit$r.squared
p_value <- summary_fit$coefficients[, 4]

png(paste0("/Users/jordanscheibe/Library/CloudStorage/OneDrive-UniversityofWyoming/Collaborations/LakeColor/LakeColor_SpatTem/Figures/Figure6/",Sys.Date(),"_Fig_6_lakecolor_vs_PC3.png"), width = 7, height=5, units = "in", res=400)
par(mar=c(3,3,1,1))
plot(lake_color$Comp.3,lake_color$avg_dwl_site, ylab = "", xlab= "",pch=16, col = lake_color$col_hex)
# add equation
text(-15, 585, 
     labels = bquote(y == .(round(summary_fit$coefficients[1,1], 2)) - .(round(abs(summary_fit$coefficients[2,1]), 2)) * x), 
     pos = 4, col = "black", cex = 1.2, adj = 0)
# Add R-squared and p-value text
text(-15, 575, 
     labels = bquote(R^2 == .(round(r_squared, 2))), 
     pos = 4, col = "black", cex = 1.2, adj = 0)

text(-15, 565, 
     labels = bquote(italic(P) == .(format.pval(p_value, digits = 2))), 
     pos = 4, col = "black", cex = 1.2, adj = 0)

abline(lm(avg_dwl_site ~  Comp.3, data=lake_color),lwd=2)
dev.off()
```

## MS stats
```{r}
summary_fit
confint(mlm, level = 0.95)
```
                 2.5 %     97.5 %
(Intercept) 533.255720 533.505075
Comp.3       -9.688173  -9.518016



