---
title: "Untitled"
author: "Jordan Von Eggers"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---


*****************************
Replace 
*****************************

# 1. Bind in all the Lagos lake data in R (stored files localled, going to call and compile and save ending (DATE_lagos_landcover_data.rds) file on Github if small enough - otherwise GDrive
```{r}
require(tidyverse)
landcover_data <- readRDS('LakeSR/1_DataPrep/1_DataDownloads/1_LAGOS/landcover_data.RDS') 

# here, I am breaking out huc12 data and removing the epanutr spatial divison landcover so we are left with huc12 only. Huc12 is likely more useful for our purposes, i.e., it will more tightly tie to individual lakes than the large regional data would. This is the data that will be included below
landcover_data_hu12 <- landcover_data |>
  filter(spatial_division != 'epanutr') |>
  rename(hu12_zoneid = zoneid) |>
  select(-spatial_division)
rm(landcover_data)


# call in other data
elev_data <- readRDS('LakeSR/1_DataPrep/1_DataDownloads/1_LAGOS/elev_data.RDS')
hydro_data <- readRDS('LakeSR/1_DataPrep/1_DataDownloads/1_LAGOS/hydro_data.RDS')
watershed_data <- readRDS('LakeSR/1_DataPrep/1_DataDownloads/1_LAGOS/watershed_data.RDS')
zoneids_data <- readRDS('LakeSR/1_DataPrep/1_DataDownloads/1_LAGOS/zoneids_data.RDS')


lake_data <-zoneids_data %>% 
        left_join(elev_data,by=join_by(lagoslakeid)) %>%
        left_join(hydro_data,by=join_by(lagoslakeid)) %>%
        left_join(watershed_data,by=join_by(lagoslakeid)) %>% 
        full_join(landcover_data_hu12,by=join_by(hu12_zoneid), multiple = "all", relationship="many-to-many")
        

#LEFT OFF HERE - add in the other geo packages we added in later on?? the soils etc


saveRDS(paste0(Sys.Date(),"_lagos_landcover_data.rds"))
```




# 2. Filter and create limnosat dataframes 
## a. Filter limnosat observations by season and LAGOS data
Filter by May 1-Oct 31 observations and keep observations with LAGOS data

```{bash}
cd /project/lakecolor/data
```

```{r}
require(tidyverse)

load("2024-07-30_full_limnosat.Rdata")
lake_data<-readRDS("2023-09-29_lagos_landcover_data.rds")


#Filter Limnosat data by observations between May 1st October 31st
new_limnosat$date_m_d <- format(new_limnosat$date, format = "%m-%d")
head(new_limnosat$date_m_d)
start_date <- format(as.Date("2023-05-01"), format = "%m-%d")
end_date <- format(as.Date("2023-10-31"), format = "%m-%d")
limnosat_may_oct <- new_limnosat[new_limnosat$date_m_d >= start_date & new_limnosat$date_m_d <= end_date, ]
start_date; end_date

nrow(limnosat_may_oct)
#42516565
nrow(new_limnosat)
#64271235
42516565/64271235
#0.6615178 - 66 % of the observations retained after filtering for Limnosat observations May 1 - Oct 31

# merge with Lagos data (should be static for each lake)
# first take out the landcover data that was attached
lagos<-lake_data[,1:15]
lagos<-distinct(lagos)
nrow(lagos)
#479950

table(unique(limnosat_may_oct$lake_nhdid)%in%unique(lagos$lake_nhdid))
 # FALSE   TRUE 
 # 42775 146301 
#  0.2923767 don't have lagos data
 
#unique number of lakes to start with
length(unique(new_limnosat$lake_nhdid))
#193667

#after filtering for may-oct season
length(unique(limnosat_may_oct$lake_nhdid))
#189076


#merge may-oct limnosat dataset with the filtered lagos dataset
limnosat_may_oct_lagos<-inner_join(limnosat_may_oct,lagos)
#Joining with `by = join_by(lake_nhdid)`

#after filtering for may-oct and observations with lagos data
length(unique(limnosat_may_oct_lagos$lake_nhdid))
#146301

nrow(limnosat_may_oct_lagos)
#36051931

36051931/64271235 #56% of observations retained after filtering for May-Oct and only those with Lagos data

#remove month/day column
limnosat_may_oct_lagos$date_m_d<-NULL

#remove all objects
rm(list=setdiff(ls(), "limnosat_may_oct_lagos"))
save.image(paste0(Sys.Date(),"_Limnosat_May_Oct_Lagos.RData"))
```


## b. Create temporal dataset

Here we are filtering by lakes that have:
1. 30 years worth of observations
2. 3 observations minimum per year
3. observations no more than 5 years apart


Filter_Limnosat_May_Oct_Lagos_For_Temporal.R
```{r}
require(tidyverse)
load("2024-07-30_Limnosat_May_Oct_Lagos.RData")

#starting number of lakes
length(unique(limnosat_may_oct_lagos$lake_nhdid))


#############################################################################
#### Filter the sites x years by only those with at least 3 observations ####
#############################################################################
observations_bySite_byYear <- limnosat_may_oct_lagos %>%
        group_by(lake_nhdid, year) %>% 
        summarize(Count = n()) 
# unqiue sites and years
nrow(observations_bySite_byYear)


morethan3obsperyear<-observations_bySite_byYear[observations_bySite_byYear$Count>=3,]
#number of unique sites and years retained
nrow(morethan3obsperyear)

#% of unique sites and years
nrow(morethan3obsperyear)/nrow(observations_bySite_byYear)

#subset the main dataframe by the years and sites that have at least 3 observations
morethan3obsperyear$lake_year<-paste(morethan3obsperyear$lake_nhdid,morethan3obsperyear$year,sep="@")
head(morethan3obsperyear)
limnosat_may_oct_lagos$lake_year<-paste(limnosat_may_oct_lagos$lake_nhdid,limnosat_may_oct_lagos$year,sep="@")
limnosat_may_oct_lagos$lake_year[1:6]

#percent of observations lost
table(limnosat_may_oct_lagos$lake_year%in%morethan3obsperyear$lake_year) 

# starting number of observations
nrow(limnosat_may_oct_lagos)

limnosat_may_oct_lagos_3obs<-limnosat_may_oct_lagos[limnosat_may_oct_lagos$lake_year%in%morethan3obsperyear$lake_year,]
# observations after removing ones with 3 observations
nrow(limnosat_may_oct_lagos_3obs)



###########################################################################
#### Filter the sites by those with 30+ years of limnosat observations #### 
###########################################################################
unique_years_bySite <- limnosat_may_oct_lagos_3obs %>%
        group_by(lake_nhdid) %>%  # Group by lake
        summarize(Number_Years = n_distinct(year))  

head(unique_years_bySite)
Yrs30<-unique_years_bySite[unique_years_bySite$Number_Years>=30,]
#number of lakes with 30_ years of data
nrow(Yrs30)


limnosat_may_oct_lagos_30yrs<-limnosat_may_oct_lagos_3obs[limnosat_may_oct_lagos_3obs$lake_nhdid%in%Yrs30$lake_nhdid,]
# should match tne number of sites
length(unique(limnosat_may_oct_lagos_30yrs$lake_nhdid))


############################################################################
#### Filter the lakes that have observations no more than 5 years apart #### 
############################################################################
limnosat_may_oct_lagos_30yrs <-limnosat_may_oct_lagos_30yrs %>% 
        arrange(lake_nhdid, year)
lake_year<-limnosat_may_oct_lagos_30yrs%>%
        select(lake_nhdid, year,date) #this grabs date too even if you don't add it here
lake_year$date<-NULL
nrow(lake_year)

#lake years after distinct
lake_year<-distinct(lake_year)
nrow(lake_year)

#number of lakes
length(unique(lake_year$lake_nhdid))



lakes_rm<-NULL
i=1
j=1
for(i in 1:length(unique(lake_year$lake_nhdid))){
        sub<-lake_year[lake_year$lake_nhdid==unique(lake_year$lake_nhdid)[i],]
        for(j in 1:(nrow(sub)-1)){
                if(sub[j+1,]$year - sub[j,]$year > 5){
                        lakes_rm<-c(lakes_rm, as.character(sub[1,1]))   
                }
        }
        print(i)
}
save.image(paste0(Sys.Date(),"_TEMPORAL_Tempfile.RData"))

limnosat_may_oct_lagos_30Yr_3ObsYr_No5Yr<-limnosat_may_oct_lagos_30yrs[limnosat_may_oct_lagos_30yrs$lake_nhdid%in%setdiff(limnosat_may_oct_lagos_30yrs$lake_nhdid,lakes_rm),]


#how many lakes?
length(unique(limnosat_may_oct_lagos_30Yr_3ObsYr_No5Yr$lake_nhdid))


length(unique(limnosat_may_oct_lagos_30Yr_3ObsYr_No5Yr$lake_nhdid))/length(unique(limnosat_may_oct_lagos$lake_nhdid))


#rename to something short and easy to work with
limnosat<-limnosat_may_oct_lagos_30Yr_3ObsYr_No5Yr
write.csv(lakes_rm,"lakes_removed_5YearsNoData.csv")
rm(list=setdiff(ls(), "limnosat"))

write_rds(limnosat,paste0(Sys.Date(),"_Limnosat_Lagos.rds"))
```
Total number of lakes = 81,875

run_Filter_Limnosat_May_Oct_Lagos_For_Temporal.sh
```{bash}
#!/bin/bash
#SBATCH --job-name Filter_Limnosat
#SBATCH --mem=60GB
#SBATCH --time=12:00:00
#SBATCH --cpus-per-task=1
#SBATCH --account=microbiome
#SBATCH --output=Filter_Limnosat_May_Oct_Lagos_For_Temporal_%A.out

cd /project/lakecolor/data/
module load arcc/1.0  gcc/12.2.0 
module load r/4.4.0

srun Rscript Filter_Limnosat_May_Oct_Lagos_For_Temporal.R
date

```



# 5. Add in climate data from PRISM

```{bash}
#load r and gdal (need gdal for terra to work)
module load arcc/1.0  gcc/12.2.0 
module load r/4.4.0
module load gdal/3.6.1
R
```

 
```{r}
# Load required packages and trim initial spatial dataset for comparison.

#Set required packages
require(tidyverse)
require(codetools)
require(terra)
require(prism)

#load in temporal data to get coordinates for each lake
limno<-readRDS("/project/lakecolor/data/2024-07-30_Limnosat_Lagos.rds")

#subset temporal to just variables of interest 
lake.coords.id <- select(limno, lake_nhdid, lake_lat_decdeg, lake_lon_decdeg)
#select kept adding a date column so I had to brute force remove it
#lake.coords.id <- lake.coords.id[-1]
#trim spatial for only one instance of each lake
lake.coords.id <- lake.coords.id %>% distinct()


# Create geometry objects from the lake coordinates

#Create a dataframe that's just the coordinates 
xy <- lake.coords.id[2:3]
#rename to lat to y and long to x
colnames(xy)[1:2] = c("y","x")
#rearrange colums so x comes first
xy <- xy[2:1]
#turn lake.coord.id into a SpatVect
lakepoints <- vect(xy, geom=c("x", "y"), crs = "epsg:4326")


# Set the prism directory (commented out is how you download PRISM data, which is a one time thing)
#set prism archive and download prism data
prism_set_dl_dir("/project/lakecolor/data/prism_input")
# get_prism_annual("ppt", 1984:2022, keepZip = FALSE)
# get_prism_annual("tmean", 1984:2022, keepZip = FALSE)
# get_prism_annual("tmax", 1984:2022, keepZip = FALSE)
# get_prism_annual("tmin", 1984:2022, keepZip = FALSE)


# Download mean, max, min, and ppt from the prism archive 

#Access prism archive (downloaded into the lake color project) for tmean/tmax/tmin/ppt annual values
tmean <- prism_archive_subset(type = "tmean", temp_period = "annual",)
tmax <- prism_archive_subset(type = "tmax", temp_period = "annual",)
tmin <- prism_archive_subset(type = "tmin", temp_period = "annual",)
ppt <- prism_archive_subset(type = "ppt", temp_period = "annual",)

# Create a raster out of the prism data 

#rasterize the data (base resolution for this function is 4km)
tmean <- rast(pd_to_file(tmean))
tmax <- rast(pd_to_file(tmax))
tmin <- rast(pd_to_file(tmin))
ppt <- rast(pd_to_file(ppt))


# Extract values from each lake point and make it into a data frame

#extract values from the data raster for each year using the location of each lake in the set (using the spatvector previously made)
tmean_from_prism <- data.frame(xy, 
                   lake.coords.id$lake_nhdid, 
                   extract(tmean, lakepoints))
tmax_from_prism <- data.frame(xy, 
                   lake.coords.id$lake_nhdid, 
                   extract(tmax, lakepoints))
tmin_from_prism <- data.frame(xy, 
                   lake.coords.id$lake_nhdid, 
                   extract(tmin, lakepoints))
ppt_from_prism <- data.frame(xy, 
                   lake.coords.id$lake_nhdid, 
                   extract(ppt, lakepoints))


# Format data 

#convert wide form to long form 
tmean_from_prism <- pivot_longer(data = tmean_from_prism, cols = starts_with("PRISM"))
tmax_from_prism <- pivot_longer(data = tmax_from_prism, cols = starts_with("PRISM"))
tmin_from_prism <- pivot_longer(data = tmin_from_prism, cols = starts_with("PRISM"))
ppt_from_prism <- pivot_longer(data = ppt_from_prism, cols = starts_with("PRISM"))
#convert string of prism data year to numeric value of year 
tmean_from_prism$name <- str_extract(tmean_from_prism$name, "\\d{4}")
tmax_from_prism$name <- str_extract(tmax_from_prism$name, "\\d{4}")
tmin_from_prism$name <- str_extract(tmin_from_prism$name, "\\d{4}")
ppt_from_prism$name <- str_extract(ppt_from_prism$name, "\\d{4}")
#rename "name" to "year" 
colnames(tmean_from_prism)[1:6] = c("lake_lon_decdeg","lake_lat_decdeg","lake_nhdid","ID","year", "tmean")
colnames(tmax_from_prism)[1:6] = c("lake_lon_decdeg","lake_lat_decdeg","lake_nhdid","ID","year", "tmax")
colnames(tmin_from_prism)[1:6] = c("lake_lon_decdeg","lake_lat_decdeg","lake_nhdid","ID","year", "tmin")
colnames(ppt_from_prism)[1:6] = c("lake_lon_decdeg","lake_lat_decdeg","lake_nhdid","ID","year", "ppt")


# Join data into one data frame 

prism_lake_data <- tmean_from_prism %>% 
  left_join(tmax_from_prism, by = c("lake_lon_decdeg","lake_lat_decdeg","lake_nhdid","ID","year")) %>% 
  left_join(tmin_from_prism, by = c("lake_lon_decdeg","lake_lat_decdeg","lake_nhdid","ID","year")) %>% 
  left_join(ppt_from_prism, by = c("lake_lon_decdeg","lake_lat_decdeg","lake_nhdid","ID","year"))

#Remove ID column
prism_lake_data$ID <- NULL

#Save the final product to the working directory 

saveRDS(prism_lake_data, file = "prism_lakepoint_data.rds")
```



# 6. Move and protect files
```{bash}
mkdir 01_Compile_Filter_Data
chmod u+w compile_data limnosat_small
mv * 01_Compile_Filter_Data/
chmod a-w *
```

# 7. Final files
The two final files from the 01_Compile_Filter_Data.Rmd and in the 01_Compile_Filter_Data folder are: 
1. 2024-07-30_Limnosat_Lagos.rds - compiled and filtered data for temporal time series, with 81,875 total lakes
2. prism_lakepoint_data.rds - PRISM data for the previous file